# Writing Functions in R {#sec-writing-functions}

```{r}
#| label: ch13-setup
#| include: false
library(tidyverse)
library(data.table)
theme_set(theme_minimal())
```

::: {.callout-note title="Learning Objectives"}
After completing this chapter, you will be able to:

- Write your own R functions using proper syntax
- Specify function arguments with default values
- Return single or multiple values from functions
- Use control flow (if/else, case_when) within functions
- Understand lexical scoping and function environments
- Apply functional programming principles with lapply and purrr
- Iterate over single and multiple inputs efficiently
- Choose between for loops and functional approaches
:::

## Why Write Your Own Functions? {#sec-why-functions}

Throughout this book, we've used many functions: some from base R [@rproject2024] (like `mean()` and `sum()`), others from packages (like `dplyr::filter()` and `ggplot2::ggplot()`) [@wickham2019tidyverse]. But the real power of programming comes from writing your own functions. (@sec-writing-simple-functions gave a first taste; here we go further.)

Functions allow you to **avoid repetition** by writing code once and using it many times. When you find yourself copying and pasting the same block of code with minor changes, that's a sign you should write a function. This also **reduces errors**—if you find a bug, you fix it in one place rather than tracking down every copy. Functions **improve readability** by letting you give meaningful names to complex operations; `calculate_gc_content(sequence)` is far clearer than the five lines of code it might contain. And functions make it easy to **share code** with colleagues or package your solutions for the broader community.

In biosciences, you'll often need custom functions for domain-specific tasks: calculating enzyme kinetics parameters, normalizing gene expression data, computing population genetics statistics, or processing experimental replicates according to your lab's specific protocols.

## Basic Function Syntax {#sec-function-syntax}

All R functions follow the same basic syntax:

```r
function_name = function(ARGUMENTS) {
  OPERATIONS
  return(VALUE)
}
```

Let's break this down:

- **function_name**: A descriptive name for your function
- **ARGUMENTS**: Input values the function accepts
- **OPERATIONS**: What the function does
- **return(VALUE)**: What the function gives back

For short functions, you can write everything on one line:

```r
my_short_func = function(x) x^2
```

::: {.callout-tip}
Give your functions short, descriptive names that indicate what they do. `calculate_gc_content` is better than `func1` or `cgc`.
:::

## A Simple Example: GC Content Calculator {#sec-gc-example}

Let's write a function that calculates the GC content of a DNA sequence --- the proportion of guanine (G) and cytosine (C) bases. (The same function is written in Python in @sec-py-functions.)

```{r}
#| label: gc-content-basic

gc_content = function(sequence) {
  # Convert to uppercase for consistency
  seq_upper = toupper(sequence)

  # Count G and C bases
  gc_count = str_count(seq_upper, "[GC]")

  # Calculate total length (excluding any non-ATGC characters)
  total_bases = str_count(seq_upper, "[ATGC]")

  # Return GC proportion
  gc_count / total_bases
}
```

Test it:

```{r}
#| label: gc-test

gc_content("ATGCGATCGATCG")
gc_content("atgcGATCgatcg")  # Works with mixed case
gc_content("AAAAAATTTTTT")   # Low GC content
gc_content("GGGGGGCCCCCC")   # High GC content
```

### Specifying Return Values {#sec-specifying-return-values}

In the example above, R automatically returned the last evaluated expression. However, it's good practice to use explicit `return()` statements, especially for complex functions:

```{r}
#| label: gc-content-explicit

gc_content = function(sequence) {
  seq_upper = toupper(sequence)
  gc_count = str_count(seq_upper, "[GC]")
  total_bases = str_count(seq_upper, "[ATGC]")
  gc_proportion = gc_count / total_bases

  return(gc_proportion)  # Explicit return
}
```

### Returning Multiple Values {#sec-returning-multiple-values}

What if we want to return more than one piece of information? Use a list or data frame:

```{r}
#| label: gc-content-detailed

gc_content_detailed = function(sequence) {
  seq_upper = toupper(sequence)

  # Count each base
  a_count = str_count(seq_upper, "A")
  t_count = str_count(seq_upper, "T")
  g_count = str_count(seq_upper, "G")
  c_count = str_count(seq_upper, "C")

  total = a_count + t_count + g_count + c_count
  gc_prop = (g_count + c_count) / total

  # Return as a data frame
  result = tibble(
    sequence = sequence,
    length = total,
    A = a_count,
    T = t_count,
    G = g_count,
    C = c_count,
    gc_content = round(gc_prop, 4)
  )

  return(result)
}

gc_content_detailed("ATGCGATCGATCGATCG")
```

### Default Argument Values {#sec-default-argument-values}

You can specify default values for arguments. This is useful when most calls will use the same value:

```{r}
#| label: gc-with-defaults

gc_content = function(sequence, digits = 4) {
  seq_upper = toupper(sequence)
  gc_count = str_count(seq_upper, "[GC]")
  total_bases = str_count(seq_upper, "[ATGC]")
  gc_proportion = gc_count / total_bases

  return(round(gc_proportion, digits))
}

gc_content("ATGCGATCGATCG")       # Uses default 4 digits
gc_content("ATGCGATCGATCG", 2)    # Override to 2 digits
```

## Environments and Lexical Scoping {#sec-lexical-scoping}

An important concept to understand is that functions operate in their own **environment**. Objects created inside a function don't appear in your global environment:

```{r}
#| label: scoping-demo

test_function = function(x) {
  internal_var = x * 2  # This stays inside the function
  return(internal_var)
}

result = test_function(5)
result

# Try to access internal_var - it doesn't exist in the global environment
# internal_var  # Would produce an error
```

This is called **lexical scoping**. R looks for objects in this order:

1. Inside the function
2. In the environment where the function was defined
3. In parent environments, up to the global environment

::: {.callout-important}
Functions are "sandboxed" --- they don't pollute your global environment with intermediate variables. This makes code cleaner and reduces the risk of naming conflicts.
:::

## Control Flow in Functions {#sec-control-flow}

### Using if/else {#sec-using-if-else}

Control flow allows your functions to behave differently based on conditions:

```{r}
#| label: enzyme-kinetics

# Calculate enzyme velocity using Michaelis-Menten kinetics
enzyme_velocity = function(substrate, vmax, km, inhibitor = NULL, ki = NULL) {

  if (is.null(inhibitor)) {
    # No inhibitor: standard Michaelis-Menten
    velocity = (vmax * substrate) / (km + substrate)
  } else {
    # Competitive inhibition
    if (is.null(ki)) {
      stop("If inhibitor concentration is provided, ki must also be specified")
    }
    apparent_km = km * (1 + inhibitor / ki)
    velocity = (vmax * substrate) / (apparent_km + substrate)
  }

  return(velocity)
}

# Test without inhibitor
enzyme_velocity(substrate = 10, vmax = 100, km = 5)

# Test with inhibitor
enzyme_velocity(substrate = 10, vmax = 100, km = 5, inhibitor = 2, ki = 3)
```

### Using case_when for Multiple Conditions {#sec-fn-case-when}

When you have many conditions, nested `if/else` statements become hard to read. Use `dplyr::case_when()` instead:

```{r}
#| label: classify-expression

# Classify gene expression fold changes
classify_expression = function(fold_change) {
  category = case_when(
    fold_change >= 2 ~ "strongly_upregulated",
    fold_change >= 1.5 ~ "upregulated",
    fold_change <= 0.5 ~ "strongly_downregulated",
    fold_change <= 0.67 ~ "downregulated",
    TRUE ~ "unchanged"
  )
  return(category)
}

# Test with a vector of fold changes
fold_changes = c(0.3, 0.6, 1.0, 1.6, 2.5)
classify_expression(fold_changes)
```

The equivalent `data.table::fcase()` function works similarly:

```{r}
#| label: classify-expression-dt

classify_expression_dt = function(fold_change) {
  fcase(
    fold_change >= 2, "strongly_upregulated",
    fold_change >= 1.5, "upregulated",
    fold_change <= 0.5, "strongly_downregulated",
    fold_change <= 0.67, "downregulated",
    default = "unchanged"
  )
}

classify_expression_dt(fold_changes)
```

## Iteration and Functional Programming {#sec-iteration}

### The Problem with for Loops {#sec-the-problem-with-for-loops}

You'll often need to apply the same operation to many inputs. The traditional approach uses `for` loops (@sec-r-for-loops):

```{r}
#| label: ch13-for-loop-example

# Calculate GC content for multiple sequences
sequences = c("ATGCGATCG", "GGGGCCCC", "AAAATTTT", "ATATATAT")

# Using a for loop
gc_results = NULL
for (i in 1:length(sequences)) {
  gc_results[i] = gc_content(sequences[i])
}
gc_results
```

This works, but has several problems:

1. **Verbose**: Lots of boilerplate code
2. **Error-prone**: Easy to make off-by-one errors
3. **Pollutes environment**: The loop variable `i` persists after the loop
4. **Not R-like**: R is designed for vectorized operations

### Functional Programming with lapply {#sec-lapply}

The **functional programming** approach treats iteration as applying a function across elements:

```{r}
#| label: lapply-example

# Much cleaner!
gc_results = lapply(sequences, gc_content)
gc_results
```

The result is a list. To get a vector instead:

```{r}
#| label: sapply-example

gc_results = sapply(sequences, gc_content)
gc_results
```

To combine results into a data frame, use our detailed function with `bind_rows()`:

```{r}
#| label: lapply-bind

lapply(sequences, gc_content_detailed) %>%
  bind_rows()
```

### Using purrr for Iteration {#sec-purrr-map}

The **purrr** package provides a consistent and powerful set of mapping functions (base R's `apply()` family, @sec-apply-family, plays the same role):

```{r}
#| label: purrr-example

# map() returns a list (like lapply)
map(sequences, gc_content)

# map_dbl() returns a numeric vector
map_dbl(sequences, gc_content)

# map_df() returns a data frame
map_df(sequences, gc_content_detailed)
```

### Creating and Reusing Named Functions {#sec-creating-and-reusing-named-functions}

For complex analyses, separate the function definition from iteration:

```{r}
#| label: named-function

# Define a function for processing protein sequences
analyze_protein = function(sequence) {
  seq_upper = toupper(sequence)
  length = nchar(seq_upper)

  # Count hydrophobic residues (simplified)
  hydrophobic = str_count(seq_upper, "[AILMFVPW]")

  # Count charged residues
  charged = str_count(seq_upper, "[DEKRH]")

  tibble(
    sequence = sequence,
    length = length,
    hydrophobic_pct = round(hydrophobic / length * 100, 1),
    charged_pct = round(charged / length * 100, 1)
  )
}

# Some sample protein sequences
proteins = c(
  "MKTAYIAKQRQISFVKSHFSRQLEERLGLIEVQAPILSRVGDGTQDNLSGAEKAVQVKVKALPDAQFEVVHSLAKWKRQQIAAALEHHHHHH",
  "MVLSPADKTNVKAAWGKVGAHAGEYGAEALERMFLSFPTTKTYFPHFDLSH",
  "GLSDGEWQQVLNVWGKVEADIAGHGQEVLIRLFTGHPETLEKFDKFKHLKTEAEMKASEDLKKHGTVVLTALGGILKKKGHHEAELKPLAQSHATKHKIPIKYLEFISDAIIHVLHSKHPGDFGADAQGAMTKALELFRNDIAAKYKELGFQG"
)

# Apply to all proteins
map_df(proteins, analyze_protein)
```

## Iterating Over Multiple Inputs {#sec-multiple-inputs}

Sometimes you need to iterate over multiple inputs simultaneously.

### Using pmap for Multiple Arguments {#sec-pmap}

```{r}
#| label: pmap-example

# Simulate growth curves with different parameters
simulate_growth = function(time, k, n0, carrying_capacity) {
  # Logistic growth model
  n = carrying_capacity / (1 + ((carrying_capacity - n0) / n0) * exp(-k * time))

  tibble(
    time = time,
    k = k,
    n0 = n0,
    K = carrying_capacity,
    population = round(n, 0)
  )
}

# Parameters for different bacterial strains
params = tibble(
  k = c(0.5, 0.8, 0.3),              # Growth rates
  n0 = c(100, 50, 200),              # Initial populations
  carrying_capacity = c(10000, 15000, 8000)  # Carrying capacities
)

# Run simulations for each strain at time = 10
pmap_df(params, ~simulate_growth(time = 10, k = ..1, n0 = ..2, carrying_capacity = ..3))
```

### Using a Data Frame of Inputs {#sec-using-a-data-frame-of-inputs}

A cleaner approach for complex cases is to pass a data frame of input combinations (@fig-df-inputs shows the result):

```{r}
#| label: fig-df-inputs
#| fig-cap: "Simulated bacterial growth curves for three strains, produced by mapping a function over a data frame of parameter combinations"

# Create all parameter combinations
simulation_params = expand.grid(
  time = seq(0, 20, by = 2),
  strain = c("wild_type", "mutant_A", "mutant_B")
) %>%
  as_tibble() %>%
  mutate(
    k = case_when(
      strain == "wild_type" ~ 0.5,
      strain == "mutant_A" ~ 0.8,
      strain == "mutant_B" ~ 0.3
    ),
    n0 = 100,
    carrying_capacity = 10000
  )

# Function that takes a row of parameters
run_simulation = function(row_data) {
  simulate_growth(
    time = row_data$time,
    k = row_data$k,
    n0 = row_data$n0,
    carrying_capacity = row_data$carrying_capacity
  ) %>%
    mutate(strain = row_data$strain)
}

# Run all simulations
results = map_df(1:nrow(simulation_params), function(i) {
  run_simulation(simulation_params[i, ])
})

# Plot the results
ggplot(results, aes(x = time, y = population, color = strain)) +
  geom_line(linewidth = 1) +
  geom_point() +
  labs(
    title = "Bacterial Growth Curves",
    x = "Time (hours)",
    y = "Population Size",
    color = "Strain"
  ) +
  scale_y_continuous(labels = scales::comma)
```

## Best Practices for Writing Functions {#sec-function-best-practices}

### 1. Use Descriptive Names {#sec-fn-descriptive-names}

```{r}
#| label: naming-examples
#| eval: false

# Good names
calculate_gc_content()
normalize_expression()
fit_growth_curve()

# Bad names
func1()
do_stuff()
f()
```

### 2. Document Your Functions {#sec-fn-document}

Add comments explaining what the function does, its arguments, and return value:

```{r}
#| label: documented-function

# Calculate the melting temperature of a DNA primer
#
# Uses the Wallace rule for short oligonucleotides (<14 bp)
# or the nearest-neighbor method for longer sequences
#
# Arguments:
#   sequence: DNA sequence (string of A, T, G, C)
#   method: "wallace" or "nearest_neighbor"
#
# Returns:
#   Melting temperature in degrees Celsius

calculate_tm = function(sequence, method = "wallace") {
  seq_upper = toupper(sequence)
  length = nchar(seq_upper)

  a_count = str_count(seq_upper, "A")
  t_count = str_count(seq_upper, "T")
  g_count = str_count(seq_upper, "G")
  c_count = str_count(seq_upper, "C")

  if (method == "wallace" || length < 14) {
    # Wallace rule: Tm = 2(A+T) + 4(G+C)
    tm = 2 * (a_count + t_count) + 4 * (g_count + c_count)
  } else {
    # Simplified nearest-neighbor approximation
    tm = 64.9 + 41 * (g_count + c_count - 16.4) / length
  }

  return(tm)
}

calculate_tm("ATGCGATCGATCG")
```

### 3. Validate Inputs {#sec-fn-validate-inputs}

Check that inputs are valid before processing:
```{r}
#| label: input-validation

gc_content_safe = function(sequence) {
  # Check input type
  if (!is.character(sequence)) {
    stop("Sequence must be a character string")
  }

  # Check for valid DNA characters
  seq_upper = toupper(sequence)
  invalid_chars = str_remove_all(seq_upper, "[ATGCN]")
  if (nchar(invalid_chars) > 0) {
    warning(paste("Invalid characters removed:", invalid_chars))
  }

  gc_count = str_count(seq_upper, "[GC]")
  total_bases = str_count(seq_upper, "[ATGC]")

  if (total_bases == 0) {
    stop("No valid DNA bases found in sequence")
  }

  return(gc_count / total_bases)
}

gc_content_safe("ATGCGATCG")     # Works fine
gc_content_safe("ATGCXYZGATCG")  # Warning about invalid characters
```

### 4. Keep Functions Focused {#sec-fn-focused}

Each function should do one thing well. Break complex operations into smaller functions:

```{r}
#| label: focused-functions

# Instead of one giant function, break it up:

# Step 1: Clean the sequence
clean_sequence = function(seq) {
  seq %>%
    toupper() %>%
    str_remove_all("[^ATGC]")
}

# Step 2: Calculate base composition
base_composition = function(seq) {
  tibble(
    A = str_count(seq, "A"),
    T = str_count(seq, "T"),
    G = str_count(seq, "G"),
    C = str_count(seq, "C")
  )
}

# Step 3: Derive statistics
sequence_stats = function(composition) {
  total = sum(composition)
  composition %>%
    mutate(
      total = total,
      gc_content = (G + C) / total,
      at_content = (A + T) / total
    )
}

# Combine them in a pipeline
"atgcGATCgatcgNNNN" %>%
  clean_sequence() %>%
  base_composition() %>%
  sequence_stats()
```

## Summary {#sec-functions-summary}

Writing functions is a fundamental programming skill that will dramatically improve your data analysis workflows:

- **Basic syntax**: `function_name = function(args) { body; return(value) }`
- **Return values**: Use explicit `return()` statements; return lists or data frames for multiple values
- **Default arguments**: Provide sensible defaults for commonly-used parameter values
- **Lexical scoping**: Functions operate in their own environment, keeping your global environment clean
- **Control flow**: Use `if/else` and `case_when()` to handle different conditions
- **Functional programming**: Prefer `lapply()`/`map()` over for loops for iteration
- **Multiple inputs**: Use `pmap()` or data frame inputs for complex parameter combinations

The next chapter (@sec-parallel-computing) will show you how to run functions in parallel, catch errors gracefully (@sec-parallel-errors), and cache results for expensive computations (@sec-caching).

::: {.callout-tip}
For quick reference on R function syntax and functional programming tools like `lapply()` and `purrr`, see @sec-appendix-r.
:::

## Additional Reading {#sec-fn-additional-reading .unnumbered}

To deepen your understanding of functions and functional programming in R:

- **[R for Data Science: Functions](https://r4ds.hadley.nz/functions)** [@wickham2023r] — Excellent introduction to writing functions
- **[Advanced R: Functions](https://adv-r.hadley.nz/functions.html)** by Hadley Wickham — Deep dive into function mechanics
- **[purrr Documentation](https://purrr.tidyverse.org/)** — Complete guide to functional programming with purrr
- **[R for Data Science: Iteration](https://r4ds.hadley.nz/iteration)** — More on map functions and iteration

For advanced function development:

- **[Advanced R: Functional Programming](https://adv-r.hadley.nz/fp.html)** — Closures, function factories, and more
- **[R Packages](https://r-pkgs.org/)** by Wickham and Bryan — When you're ready to package your functions

## Exercises {#sec-fn-exercises .unnumbered}

::: {.callout-tip title="Practice Exercises"}
**Exercise 1: Basic Function**

Write a function called `count_codons()` that takes a DNA sequence and returns the number of complete codons (groups of 3 nucleotides). For example, "ATGCGA" has 2 complete codons.

**Exercise 2: Function with Multiple Returns**

Write a function `analyze_sequence()` that takes a DNA sequence and returns a data frame with:
- The sequence length
- GC content
- Number of each base (A, T, G, C)
- Whether it starts with a start codon (ATG)
- Whether it ends with a stop codon (TAA, TAG, or TGA)

**Exercise 3: Control Flow**

Write a function `translate_codon()` that takes a three-letter codon and returns the corresponding amino acid (use a simplified table with just 5-10 codons). Include error handling for invalid inputs.

**Exercise 4: Iteration**

You have a vector of 10 DNA sequences. Use `map_df()` to apply your `analyze_sequence()` function from Exercise 2 to all of them and combine the results into a single data frame.

**Exercise 5: Multiple Inputs**

Write a function `dilution_series()` that takes an initial concentration and dilution factor, then returns a data frame showing the concentration at each step for a specified number of dilutions. Use `pmap()` to apply this to multiple starting conditions.

**Exercise 6: Refactoring**

Take the following for loop and rewrite it using functional programming:

```r
results = c()
for (i in 1:length(sequences)) {
  gc = gc_content(sequences[i])
  if (gc > 0.5) {
    results = c(results, "high")
  } else {
    results = c(results, "low")
  }
}
```
:::
