# Parallel Computing in R {#sec-parallel-computing}

```{r}
#| label: ch14-setup
#| include: false
#| cache: false
#| message: false
#| warning: false

if (!require("pacman")) install.packages("pacman")
pacman::p_load(tictoc, parallel, future, future.apply, tidyverse, furrr)
theme_set(theme_minimal())

## Set future resolution strategy
plan(multisession)
```

::: {.callout-note title="Learning Objectives"}
After completing this chapter, you will be able to:

- Understand when parallel computing provides performance benefits
- Use the future ecosystem for parallel programming in R
- Apply future_lapply() and future_map() for parallel iteration
- Choose between forking and socket-based parallelization
- Understand the differences between explicit and implicit parallelization
- Avoid common pitfalls like resource competition
- Debug functions and handle errors in parallel contexts
- Cache results to avoid recomputation
:::

## Why Parallel Computing? {#sec-why-parallel}

Modern computers have multiple processor cores, but by default R only uses one. This means that when you run a computationally intensive analysis --- like bootstrapping, permutation tests, or processing thousands of samples --- most of your computer's power sits idle.

**Parallel computing** lets you use all available cores simultaneously, potentially reducing computation time by a factor equal to your core count. A task that takes 8 hours on one core might take just 1 hour across 8 cores.

In biosciences, parallel computing is especially valuable for:

- **Bootstrapping** for confidence intervals
- **Permutation tests** for significance testing
- **Simulation studies** (e.g., population genetics)
- **Processing many samples** (e.g., analyzing hundreds of RNA-seq files)
- **Cross-validation** in machine learning
- **Parameter sweeps** for computational models

## How Many Cores Do You Have? {#sec-cores}

Before parallelizing, find out how many cores are available:

```{r}
#| label: detect-cores
#| cache: false

parallel::detectCores()
```

This number includes both physical cores and "logical" cores (from technologies like Intel's hyperthreading). For CPU-intensive tasks, your effective speedup is typically limited by the number of physical cores.

## A Motivating Example {#sec-parallel-example}

Let's start with a simple example to demonstrate the power of parallelization. We'll create a deliberately slow function that simulates a computationally expensive operation:

```{r}
#| label: slow-function

# Simulate analyzing a biological sample (takes 2 seconds each)
analyze_sample = function(sample_id) {
  # Simulate processing time
  Sys.sleep(2)

  # Return some fake results
  tibble(
    sample_id = sample_id,
    gene_count = sample(5000:20000, 1),
    quality_score = runif(1, 0.8, 1.0)
  )
}
```

### Serial Execution {#sec-serial-execution}

First, let's process 8 samples the traditional way:

```{r}
#| label: serial-example
#| cache: true

library(tictoc)

tic()
serial_results = lapply(1:8, analyze_sample) %>% bind_rows()
toc()
```

As expected, processing 8 samples at 2 seconds each takes about 16 seconds.

### Parallel Execution {#sec-parallel-execution}

Now let's do the same thing in parallel using the **future.apply** package:

```{r}
#| label: parallel-example
#| cache: true

library(future.apply)
plan(multisession)  # Enable parallel processing

tic()
parallel_results = future_lapply(1:8, analyze_sample) %>% bind_rows()
toc()
```

The parallel version is dramatically faster! With 8 cores, we can process all 8 samples simultaneously, reducing the time from ~16 seconds to ~2 seconds.

Let's verify the results are identical:

```{r}
#| label: compare-results

# Results have same structure (values differ due to random generation)
all.equal(names(serial_results), names(parallel_results))
```

## The Future Ecosystem {#sec-future}

The examples above use the **future** package ecosystem, which provides a unified, elegant approach to parallel programming in R. The key packages are listed in @tbl-future-packages:

| Package | Description |
|:--------|:------------|
| **future** | Core package defining how code is evaluated |
| **future.apply** | Parallel versions of apply functions |
| **furrr** | Parallel versions of purrr functions |

: The future ecosystem {#tbl-future-packages}

::: {.callout-note title="Base R's `parallel` package"}
Older code, and many package internals, use the **parallel** package that ships with R rather than **future**. The pattern is: start a set of worker processes, hand them work with a parallel `lapply()`, and shut them down:

```{r}
#| label: base-parallel
#| eval: false

library(parallel)
n_cores <- detectCores() - 1          # leave one core for the OS
cl <- makeCluster(n_cores)            # start the workers
results <- parLapply(cl, data_list, analysis_function)
stopCluster(cl)                       # always shut them down
```

`future`'s `plan(multisession)` does the same thing behind the scenes, but it also manages the workers for you, ships the variables and packages your function needs, and handles random seeds correctly. You should be able to read the `parallel` version when you meet it; write new code with `future` and `furrr`.
:::

### Setting Up Parallel Execution {#sec-setting-up-parallel-execution}

Before running parallel code, you must set a "plan" that determines how futures are resolved:

```{r}
#| label: plan-options
#| eval: false

library(future)

# Sequential (no parallelization - useful for debugging)
plan(sequential)

# Parallel using multiple R sessions (works on all systems)
plan(multisession)

# Parallel using forked processes (faster, but Unix/Mac only)
plan(multicore)
```

For most users, `plan(multisession)` is the safest choice as it works on all operating systems.

### future_lapply: Parallel Apply {#sec-future-lapply}

The `future_lapply()` function is a drop-in replacement for `lapply()` (@sec-lapply):

```{r}
#| label: future-lapply-demo
#| cache: true

# Serial version
# lapply(1:4, function(x) x^2)

# Parallel version - just add "future_"
future_lapply(1:4, function(x) x^2)
```

### furrr: Parallel purrr {#sec-furrr}

If you prefer **purrr**'s syntax (@sec-purrr-map), **furrr** provides parallel versions of all map functions:

```{r}
#| label: furrr-demo
#| cache: true

library(furrr)

# Serial: map_dbl(1:4, function(x) x^2)
# Parallel:
future_map_dbl(1:4, function(x) x^2)

# Returns a data frame directly
future_map_dfr(1:4, function(x) {
  tibble(value = x, squared = x^2)
})
```

## A Real Example: Bootstrapping Gene Expression {#sec-bootstrap-example}

Let's apply parallel computing to a realistic bioinformatics task: bootstrapping to estimate confidence intervals for differential expression.

### Setup {#sec-par-setup}

```{r}
#| label: bootstrap-setup

set.seed(1234)

# Simulate gene expression data for 1000 genes, 50 samples
n_genes = 1000
n_samples = 50

expression_data = tibble(
  gene_id = rep(paste0("gene_", 1:n_genes), each = n_samples),
  sample_id = rep(paste0("sample_", 1:n_samples), n_genes),
  group = rep(rep(c("control", "treatment"), each = n_samples/2), n_genes),
  expression = rnorm(n_genes * n_samples, mean = 10, sd = 2)
) %>%
  # Add a treatment effect to some genes
  mutate(
    expression = if_else(
      gene_id %in% paste0("gene_", 1:100) & group == "treatment",
      expression + rnorm(n(), 2, 0.5),  # Upregulated genes
      expression
    )
  )

head(expression_data)
```

### Bootstrap Function {#sec-bootstrap-function}

```{r}
#| label: bootstrap-function

# Function to bootstrap fold change for one gene
bootstrap_fold_change = function(gene_data) {
  # Resample within each group
  boot_sample = gene_data %>%
    group_by(group) %>%
    slice_sample(prop = 1, replace = TRUE) %>%
    ungroup()

  # Calculate mean expression per group
  means = boot_sample %>%
    group_by(group) %>%
    summarise(mean_expr = mean(expression), .groups = "drop")

  # Calculate fold change (treatment / control)
  fc = means$mean_expr[means$group == "treatment"] /
       means$mean_expr[means$group == "control"]

  return(fc)
}

# Function to run many bootstrap iterations for one gene
run_bootstrap = function(gene_id, data, n_boot = 1000) {
  gene_data = data %>% filter(gene_id == !!gene_id)

  # Run bootstrap iterations
  boot_fcs = replicate(n_boot, bootstrap_fold_change(gene_data))

  # Return summary statistics
  tibble(
    gene_id = gene_id,
    mean_fc = mean(boot_fcs),
    lower_ci = quantile(boot_fcs, 0.025),
    upper_ci = quantile(boot_fcs, 0.975)
  )
}
```

### Compare Serial vs Parallel {#sec-compare-serial-vs-parallel}

Let's bootstrap confidence intervals for 20 genes:

```{r}
#| label: bootstrap-comparison
#| cache: true

genes_to_test = paste0("gene_", 1:20)

# Serial version
tic()
serial_boot = map_dfr(genes_to_test, run_bootstrap,
                      data = expression_data, n_boot = 500)
toc()

# Parallel version
tic()
parallel_boot = future_map_dfr(genes_to_test, run_bootstrap,
                                data = expression_data, n_boot = 500,
                                .options = furrr_options(seed = TRUE))
toc()
```

The parallel version should be substantially faster, with speedup proportional to your core count.

### View Results {#sec-view-results}

```{r}
#| label: bootstrap-results

parallel_boot %>%
  mutate(
    significant = lower_ci > 1 | upper_ci < 1,
    direction = case_when(
      lower_ci > 1 ~ "upregulated",
      upper_ci < 1 ~ "downregulated",
      TRUE ~ "not significant"
    )
  ) %>%
  head(10)
```

## Forking vs Sockets {#sec-fork-socket}

There are two main approaches to parallelization:

### Socket-based (multisession) {#sec-multisession}

- Creates new R sessions for each worker
- Works on **all operating systems** including Windows
- Requires copying data to each worker (slower startup)
- Selected with `plan(multisession)`

### Fork-based (multicore) {#sec-multicore}

- Clones the current R session
- Works on **Unix/Mac only** (not Windows)
- Shares memory with parent process (faster)
- Selected with `plan(multicore)`
- May cause issues inside an IDE console (RStudio in particular); run from the terminal for best results

```{r}
#| label: plan-comparison
#| eval: false

# Safe choice for all systems
plan(multisession)

# Faster on Unix/Mac (run scripts from terminal)
plan(multicore)

# Specify number of workers
plan(multisession, workers = 4)
```

::: {.callout-warning}
On Windows, `plan(multicore)` silently falls back to `plan(sequential)`. Always use `plan(multisession)` for cross-platform code.
:::

## Handling Errors in Parallel Code {#sec-parallel-errors}

Errors in parallel code can be tricky to debug. The **purrr** package provides `safely()` and `possibly()` to handle errors gracefully:

```{r}
#| label: error-handling

# A function that sometimes fails
risky_analysis = function(x) {
  if (x == 5) stop("Error processing sample 5!")
  tibble(sample = x, result = x^2)
}

# Wrap in safely() to catch errors
safe_analysis = safely(risky_analysis, otherwise = NULL)

# Run in parallel - won't crash if one fails
results = future_map(1:10, safe_analysis)

# Extract successful results
successful = map(results, "result") %>%
  compact() %>%  # Remove NULLs

  bind_rows()

# See which failed
errors = map(results, "error") %>%
  compact()

cat("Successful:", nrow(successful), "\n")
cat("Failed:", length(errors), "\n")
```

For simpler error handling, use `possibly()`:

```{r}
#| label: possibly-example

# Returns NA for failed cases
robust_analysis = possibly(risky_analysis, otherwise = tibble(sample = NA, result = NA))

future_map_dfr(1:10, robust_analysis)
```

## Caching Results {#sec-caching}

For expensive computations, cache intermediate results so you don't have to recompute them:

```{r}
#| label: memoise-example
#| cache: true

library(memoise)

# Create a slow function
slow_gc_analysis = function(sequence) {
  Sys.sleep(1)  # Simulate slow computation
  tibble(
    sequence = sequence,
    gc_content = str_count(toupper(sequence), "[GC]") / nchar(sequence)
  )
}

# Create memoised version
memo_gc_analysis = memoise(slow_gc_analysis)

# First call - slow
tic()
result1 = memo_gc_analysis("ATGCGATCGATCG")
toc()

# Second call with same input - instant (cached)
tic()
result2 = memo_gc_analysis("ATGCGATCGATCG")
toc()
```

### Persistent Caching {#sec-persistent-caching}

To cache results across R sessions (useful for long analyses):

```{r}
#| label: persistent-cache
#| eval: false

# Create a cache directory
cache_dir = "analysis_cache"
if (!dir.exists(cache_dir)) dir.create(cache_dir)

# Create memoised function with persistent cache
memo_gc_persistent = memoise(
  slow_gc_analysis,
  cache = cache_filesystem(cache_dir)
)

# Results will be saved to disk and available in future sessions
```

## When to Use Parallel Computing {#sec-when-parallel}

Parallel computing isn't always faster. Consider:

### Good Candidates for Parallelization {#sec-good-candidates-for-parallelization}

- **Independent iterations**: Each iteration doesn't depend on others
- **CPU-intensive tasks**: Complex calculations, simulations
- **Many iterations**: Overhead is amortized across many runs
- **Long-running operations**: Minutes or hours of total computation

### Poor Candidates {#sec-poor-candidates}

- **Fast operations**: Overhead exceeds computation time
- **Few iterations**: Not enough work to distribute
- **Memory-intensive**: Each worker needs its own copy of data
- **I/O bound**: Waiting for disk/network, not CPU

### Amdahl's Law {#sec-amdahls-law}

The maximum speedup from parallelization is limited by the sequential portion of your code. If 50% of your code must run sequentially, you can never achieve more than 2× speedup, regardless of core count (@fig-amdahl).

```{r}
#| label: fig-amdahl
#| fig-cap: "Amdahl's law: speedup as a function of core count for different parallel fractions"
#| echo: false

expand.grid(
  cores = 2^(0:7),
  parallel_fraction = c(0.5, 0.75, 0.9, 0.95, 0.99)
) %>%
  mutate(
    speedup = 1 / (1 - parallel_fraction + parallel_fraction/cores),
    parallel_pct = scales::percent(parallel_fraction, accuracy = 1)
  ) %>%
  ggplot(aes(x = cores, y = speedup, color = parallel_pct)) +
  geom_line(linewidth = 1) +
  scale_x_continuous(trans = "log2", breaks = 2^(0:7)) +
  labs(
    title = "Amdahl's Law: Diminishing Returns from Parallelization",
    subtitle = "Speedup is limited by the fraction of code that can run in parallel",
    x = "Number of Cores",
    y = "Speedup Factor",
    color = "Parallel\nFraction"
  )
```

## Best Practices {#sec-parallel-best-practices}

### 1. Start Sequential, Then Parallelize {#sec-par-start-sequential}

Always develop and debug your code in sequential mode first:

```{r}
#| label: debug-sequential
#| eval: false

# Debug with sequential execution
plan(sequential)

# Test with a small subset
test_results = map_dfr(1:3, my_analysis_function)

# Once working, switch to parallel
plan(multisession)
full_results = future_map_dfr(1:1000, my_analysis_function)
```

### 2. Set Random Seeds {#sec-par-random-seeds}

For reproducible parallel simulations:

```{r}
#| label: parallel-seeds
#| eval: false

# future.apply
future_lapply(1:10, my_function, future.seed = 123)

# furrr
future_map(1:10, my_function, .options = furrr_options(seed = 123))
```

### 3. Monitor Progress {#sec-par-monitor-progress}

Use progress bars to track long-running parallel jobs:

```{r}
#| label: progress-bar
#| eval: false

library(progressr)

handlers("txtprogressbar")

with_progress({
  p = progressor(steps = 100)

  results = future_map(1:100, function(x) {
    p()  # Signal progress
    slow_function(x)
  })
})
```

### 4. Manage Memory {#sec-par-manage-memory}

Each parallel worker needs memory for:
- A copy of R
- Any data passed to the function
- Results being computed

If you run out of memory, reduce the number of workers:

```{r}
#| label: reduce-workers
#| eval: false

# Use fewer workers for memory-intensive tasks
plan(multisession, workers = 2)
```

### 5. Know When to Leave Your Laptop {#sec-parallel-to-cluster}

Everything in this chapter parallelizes *within one machine*: `plan(multisession, workers = 8)` cannot use more cores than your laptop has, and every worker shares its memory. When a job needs more cores, more memory, or more hours than that, the next step is a cluster (@sec-hpc-talapas), where the SLURM scheduler [@yoo2003slurm] hands your script a slice of a compute node. The R code needs almost no change, but two habits matter:

- **Ask `future` for the cores SLURM gave you, not the cores the node has.** On a shared node `parallel::detectCores()` reports the whole machine (often 128 cores), so use the allocation instead (see @tbl-slurm-env-vars): `plan(multisession, workers = as.integer(Sys.getenv("SLURM_CPUS_PER_TASK", "1")))`. Use `multisession`, not `multicore`, inside a batch job unless you have verified forking works with your packages.
- **Scale *out* with job arrays rather than *up* with more workers.** Many independent samples are better submitted as a SLURM job array (@sec-hpc-job-arrays), one R process per sample, each with its own memory limit and log, than as one giant `future_map()` on one node. To make that R process reproducible on any machine, package its dependencies with `renv` or run it inside an Apptainer container [@kurtzer2017singularity], the cluster-safe counterpart of Docker [@merkel2014docker].

## Summary {#sec-parallel-summary}

Parallel computing can dramatically speed up computationally intensive analyses:

- The **future** ecosystem provides a simple, unified approach to parallelization
- Use `plan(multisession)` for cross-platform compatibility
- Replace `lapply()` with `future_lapply()` or `map()` with `future_map()`
- Always develop in sequential mode first, then parallelize
- Use `safely()` or `possibly()` to handle errors gracefully
- Cache expensive results with **memoise** for persistent storage
- Consider Amdahl's law: not everything benefits from parallelization

Parallel programming in R has never been easier. With just a few lines of code, you can harness all your computer's cores for faster, more efficient analyses.

::: {.callout-tip}
For reference on R functions used in parallel computing workflows, see @sec-appendix-r.
:::

## Exercises {#sec-par-exercises .unnumbered}

::: {.callout-tip title="Practice Exercises"}
**Exercise 1: Basic Parallelization**

Write a function that simulates processing a biological sample (include a `Sys.sleep(1)` to simulate processing time). Use `future_lapply()` to process 10 samples in parallel and compare the time to serial execution.

**Exercise 2: Bootstrap Analysis**

Using the gene expression simulation from this chapter:
1. Write a function to bootstrap the mean expression of a single gene
2. Apply it in parallel to 50 genes
3. Calculate 95% confidence intervals for each gene

**Exercise 3: Error Handling**

Create a function that analyzes DNA sequences but fails if the sequence contains invalid characters. Use `safely()` to process a list of sequences where some are invalid, and report which ones failed.

**Exercise 4: Parameter Sweep**

Write a function that simulates bacterial growth with parameters for growth rate (k), carrying capacity (K), and initial population (n0). Use parallel computing to run simulations across a grid of parameter combinations.

**Exercise 5: Caching**

Create a memoised version of a "slow" function that calculates protein properties. Demonstrate that repeated calls with the same input are instant.

**Exercise 6: Benchmarking**

For a computationally intensive task of your choice:
1. Time the serial version
2. Time the parallel version with 2, 4, and 8 workers
3. Calculate the speedup factor for each
4. Does the speedup match what you'd expect from Amdahl's law?
:::

## Additional Reading {#sec-par-additional-reading .unnumbered}

- [future package documentation](https://future.futureverse.org/)
- [furrr package documentation](https://furrr.futureverse.org/)
- [Parallel Computing with R: A Brief Review](https://arxiv.org/abs/1912.11144) by Dirk Eddelbuettel
- [R for Data Science: Iteration](https://r4ds.hadley.nz/iteration.html) chapter on functional programming
