# LaTeX for Scientific Documents {#sec-latex-basics}

```{r}
#| label: ch16-setup
#| include: false
library(tidyverse)
library(gt)
theme_set(theme_minimal())
```

::: {.callout-note title="Learning Objectives"}
After completing this chapter, you will be able to:

- Explain what LaTeX is and why it's valuable for scientific writing
- Use LaTeX syntax for mathematical notation in Quarto documents
- Write inline and display equations
- Format common mathematical operators, Greek letters, and symbols
- Create matrices and complex mathematical expressions
- Understand the relationship between LaTeX, Markdown, and Quarto
:::

## What is LaTeX? {#sec-what-is-latex}

**LaTeX** (pronounced "Lah-tech" or "Lay-tech"—to rhyme with "Bertolt Brecht") is a document preparation system for high-quality typesetting. Created by Leslie Lamport in 1984, it builds on Donald Knuth's TeX typesetting system and has become the standard for scientific and technical documentation [@lamport1994latex]. Quarto (@sec-quarto-documents) uses LaTeX under the hood whenever it renders to PDF [@quarto2024].

### Why LaTeX Matters {#sec-why-latex-matters}

LaTeX excels in several key areas:

**Precision in Mathematics**
:   LaTeX produces beautifully typeset mathematical equations that are impossible to achieve in standard word processors.

**Consistent Formatting**
:   Document structure is controlled by logical commands, ensuring consistent formatting throughout.

**Large Document Management**
:   Cross-references, bibliographies, tables of contents, and indexes are handled automatically.

**Multi-lingual Support**
:   LaTeX handles complex scripts and languages with ease.

**Professional Output**
:   LaTeX is the standard for journal articles, theses, and technical books in mathematics, physics, computer science, and engineering.

::: {.callout-tip}
You don't need to learn full LaTeX to benefit from it! Quarto and R Markdown let you use LaTeX syntax for mathematical notation while handling document layout automatically (see @sec-math-equations).
:::

### LaTeX Resources {#sec-latex-resources}

- [The LaTeX Project](https://www.latex-project.org) — Official documentation and resources
- [Overleaf Documentation](https://www.overleaf.com/learn) — Excellent LaTeX tutorials
- [LaTeX Wikibook](https://en.wikibooks.org/wiki/LaTeX) — Comprehensive community guide

## LaTeX in Quarto Documents {#sec-latex-in-quarto}

One of LaTeX's greatest strengths is its seamless integration into Quarto and R Markdown documents. You can include mathematical expressions directly in your text without needing to learn the full LaTeX document system.

### Inline vs. Display Equations {#sec-inline-display}

There are two ways to include mathematical expressions:

**Inline equations** appear within the text flow, surrounded by single dollar signs:

```markdown
This equation, $y = \frac{1}{2}$, is included inline with the text.
```

Renders as: This equation, $y = \frac{1}{2}$, is included inline with the text.

**Display equations** appear on their own line, centered, using double dollar signs:

```markdown
This equation is displayed on a separate line:

$$y = \frac{1}{2}$$
```

Renders as: This equation is displayed on a separate line:

$$y = \frac{1}{2}$$

::: {.callout-note}
Use inline equations for simple expressions mentioned in text. Use display equations for important formulas that deserve visual emphasis or are too complex to fit inline.
:::

## Basic Mathematical Notation {#sec-basic-math}

### Operators and Arithmetic {#sec-operators}

LaTeX provides commands for mathematical operations:

| Expression | Code | Description |
|:-----------|:-----|:------------|
| $a + b$ | `$a + b$` | Addition |
| $a - b$ | `$a - b$` | Subtraction |
| $a \times b$ | `$a \times b$` | Multiplication |
| $a \div b$ | `$a \div b$` | Division |
| $\frac{a}{b}$ | `$\frac{a}{b}$` | Fraction |
| $a^n$ | `$a^n$` | Superscript (power) |
| $a_n$ | `$a_n$` | Subscript |
| $a^{n+1}$ | `$a^{n+1}$` | Multi-character superscript |
| $a_{i,j}$ | `$a_{i,j}$` | Multi-character subscript |

: Basic arithmetic operators {#tbl-arithmetic}

### Roots and Powers {#sec-roots}

| Expression | Code | Description |
|:-----------|:-----|:------------|
| $\sqrt{x}$ | `$\sqrt{x}$` | Square root |
| $\sqrt[n]{x}$ | `$\sqrt[n]{x}$` | nth root |
| $x^2$ | `$x^2$` | Square |
| $e^x$ | `$e^x$` | Exponential |

: Roots and powers {#tbl-roots}

Example with multiple operations:

```markdown
$$ \large a^x, \sqrt[n]{x}, \vec{\jmath}, \tilde{\imath} $$
```

$$ \large a^x, \sqrt[n]{x}, \vec{\jmath}, \tilde{\imath} $$

## Greek Letters {#sec-greek-letters}

Greek letters are essential in scientific notation. LaTeX provides commands for all Greek letters (@tbl-greek-lower and @tbl-greek-upper):

### Lowercase Greek Letters {#sec-lowercase-greek-letters}

| Letter | Code | Letter | Code | Letter | Code |
|:-------|:-----|:-------|:-----|:-------|:-----|
| $\alpha$ | `\alpha` | $\iota$ | `\iota` | $\rho$ | `\rho` |
| $\beta$ | `\beta` | $\kappa$ | `\kappa` | $\sigma$ | `\sigma` |
| $\gamma$ | `\gamma` | $\lambda$ | `\lambda` | $\tau$ | `\tau` |
| $\delta$ | `\delta` | $\mu$ | `\mu` | $\upsilon$ | `\upsilon` |
| $\epsilon$ | `\epsilon` | $\nu$ | `\nu` | $\phi$ | `\phi` |
| $\zeta$ | `\zeta` | $\xi$ | `\xi` | $\chi$ | `\chi` |
| $\eta$ | `\eta` | $\pi$ | `\pi` | $\psi$ | `\psi` |
| $\theta$ | `\theta` | $\omega$ | `\omega` | | |

: Lowercase Greek letters {#tbl-greek-lower}

Example:

```markdown
$$ \large \alpha, \beta, \gamma $$
```

$$ \large \alpha, \beta, \gamma $$

### Uppercase Greek Letters {#sec-uppercase-greek-letters}

| Letter | Code | Letter | Code | Letter | Code |
|:-------|:-----|:-------|:-----|:-------|:-----|
| $\Gamma$ | `\Gamma` | $\Lambda$ | `\Lambda` | $\Phi$ | `\Phi` |
| $\Delta$ | `\Delta` | $\Xi$ | `\Xi` | $\Psi$ | `\Psi` |
| $\Theta$ | `\Theta` | $\Pi$ | `\Pi` | $\Omega$ | `\Omega` |
| $\Sigma$ | `\Sigma` | $\Upsilon$ | `\Upsilon` | | |

: Uppercase Greek letters {#tbl-greek-upper}

::: {.callout-tip}
Many uppercase Greek letters look identical to Latin letters (A, B, E, H, I, K, M, N, O, P, T, X, Z) and don't have separate LaTeX commands—just use the Latin letter. Several lowercase letters have a second form (`\varepsilon` $\varepsilon$, `\varphi` $\varphi$, `\vartheta` $\vartheta$), and a bold Greek letter needs `\boldsymbol{\beta}` rather than `\mathbf`. The names and pronunciations of all 24 letters are listed in @tbl-greek-names.
:::

::: {.callout-note title="Greek letters in R plot labels"}
LaTeX works in Quarto text, not inside an R plot. For an axis label or title that needs a Greek letter, use R's `expression()` (or the `latex2exp` package, which translates LaTeX strings):

```{r}
#| label: greek-in-r
#| eval: false
ggplot(df, aes(x, y)) +
  geom_point() +
  labs(x = expression(mu), y = expression(sigma^2),
       title = expression(paste("Mean = ", mu, ", SD = ", sigma)))
```

In matplotlib the label string itself can hold LaTeX: `plt.xlabel(r"$\mu$")`.
:::

## Comparison and Relation Operators {#sec-comparison}

| Expression | Code | Description |
|:-----------|:-----|:------------|
| $=$ | `$=$` | Equal to |
| $\neq$ | `$\neq$` | Not equal to |
| $<$ | `$<$` | Less than |
| $>$ | `$>$` | Greater than |
| $\leq$ | `$\leq$` | Less than or equal |
| $\geq$ | `$\geq$` | Greater than or equal |
| $\approx$ | `$\approx$` | Approximately equal |
| $\sim$ | `$\sim$` | Similar to / distributed as |
| $\nsim$ | `$\nsim$` | Not similar to |
| $\equiv$ | `$\equiv$` | Equivalent / identical |
| $\propto$ | `$\propto$` | Proportional to |
| $\in$ | `$\in$` | Element of |
| $\notin$ | `$\notin$` | Not element of |
| $\subset$ | `$\subset$` | Subset of |
| $\subseteq$ | `$\subseteq$` | Subset or equal |

: Comparison and relation operators {#tbl-comparison}

Example:

```markdown
$$ \large \approx, \neq, \nsim $$
```

$$ \large \approx, \neq, \nsim $$

## Special Symbols {#sec-special-symbols}

LaTeX provides a wide range of special mathematical symbols:

| Expression | Code | Description |
|:-----------|:-----|:------------|
| $\partial$ | `$\partial$` | Partial derivative |
| $\infty$ | `$\infty$` | Infinity |
| $\nabla$ | `$\nabla$` | Del / gradient |
| $\mathbb{R}$ | `$\mathbb{R}$` | Real numbers |
| $\mathbb{N}$ | `$\mathbb{N}$` | Natural numbers |
| $\mathbb{Z}$ | `$\mathbb{Z}$` | Integers |
| $\forall$ | `$\forall$` | For all |
| $\exists$ | `$\exists$` | There exists |
| $\emptyset$ | `$\emptyset$` | Empty set |
| $\cup$ | `$\cup$` | Union |
| $\cap$ | `$\cap$` | Intersection |
| $\to$ | `$\to$` | Arrow (also `\rightarrow`, `\leftarrow`) |
| $\Rightarrow$ | `$\Rightarrow$` | Implies (`\Leftrightarrow` for "if and only if") |
| $\pm$ | `$\pm$` | Plus or minus |
| $\cdot$ | `$\cdot$` | Centered dot |
| $\ldots$ | `$\ldots$` | Baseline dots ($x_1, \ldots, x_n$) |
| $\cdots$ | `$\cdots$` | Horizontal dots ($x_1 + \cdots + x_n$) |
| $\vdots$ | `$\vdots$` | Vertical dots |
| $\ddots$ | `$\ddots$` | Diagonal dots |

: Special mathematical symbols {#tbl-special-symbols}

Logic symbols, arrows, number sets, and many more are tabulated in @sec-appendix-latex.

Example:

```markdown
$$ \large \partial, \mathbb{R}, \infty $$
```

$$ \large \partial, \mathbb{R}, \infty $$

## Sums, Products, and Integrals {#sec-sums-integrals}

### Summation and Products {#sec-summation-and-products}

| Expression | Code | Description |
|:-----------|:-----|:------------|
| $\sum_{i=1}^{n} x_i$ | `$\sum_{i=1}^{n} x_i$` | Summation |
| $\prod_{i=1}^{n} x_i$ | `$\prod_{i=1}^{n} x_i$` | Product |
| $\bigcup_{i=1}^{n} A_i$ | `$\bigcup_{i=1}^{n} A_i$` | Union |
| $\bigcap_{i=1}^{n} A_i$ | `$\bigcap_{i=1}^{n} A_i$` | Intersection |

: Summation and product notation {#tbl-sums}

### Integrals {#sec-integrals}

| Expression | Code | Description |
|:-----------|:-----|:------------|
| $\int f(x) \, dx$ | `$\int f(x) \, dx$` | Integral |
| $\int_a^b f(x) \, dx$ | `$\int_a^b f(x) \, dx$` | Definite integral |
| $\iint f(x,y) \, dx \, dy$ | `$\iint f(x,y) \, dx \, dy$` | Double integral |
| $\iiint$ | `$\iiint$` | Triple integral |
| $\oint$ | `$\oint$` | Contour integral |

: Integral notation {#tbl-integrals}

### Example: Integral Equation {#sec-example-integral-equation}

```markdown
$$\iint xy^2\,dx\,dy = \frac{1}{6}x^2y^3$$
```

$$\iint xy^2\,dx\,dy = \frac{1}{6}x^2y^3$$

## Statistical and Probability Equations {#sec-stat-equations}

Scientific writing often requires statistical formulas. Here are common examples:

### Binomial Distribution {#sec-binomial-distribution}

The probability mass function for binomial sampling:

```markdown
$$\large f(k) = {n \choose k} p^{k} (1-p)^{n-k}$$
```

$$\large f(k) = {n \choose k} p^{k} (1-p)^{n-k}$$

The binomial coefficient can equally be written `\binom{n}{k}`, which reads more clearly in source.

### Poisson Distribution {#sec-poisson-distribution}

The probability mass function:

```markdown
$$\large P(Y=r) = \frac{e^{-\mu}\mu^r}{r!}$$
```

$$\large P(Y=r) = \frac{e^{-\mu}\mu^r}{r!}$$

::: {.callout-tip}
Multi-letter names in math mode are typeset as a product of italic variables (`Pr` becomes $Pr$, i.e. $P \times r$). Use `\Pr`, `\text{Var}`, `\operatorname{Cov}`, or `\mathrm{SE}` for upright function names: $\Pr(Y = r)$, $\text{Var}(X)$.
:::

### Normal Distribution {#sec-normal-distribution}

The probability density function:

$$f(x) = \frac{1}{\sigma\sqrt{2\pi}} e^{-\frac{1}{2}\left(\frac{x-\mu}{\sigma}\right)^2}$$

### Sample Mean and Variance {#sec-sample-mean-and-variance}

Mean: $\bar{x} = \frac{1}{n}\sum_{i=1}^{n} x_i$

Variance: $s^2 = \frac{1}{n-1}\sum_{i=1}^{n}(x_i - \bar{x})^2$

## Matrices {#sec-latex-matrix-basics}

LaTeX provides several environments for typesetting matrices:

### Basic Matrix {#sec-basic-matrix}

Use the `matrix` environment for a matrix without brackets:

```markdown
$$\begin{matrix}
a & b \\
c & d
\end{matrix}$$
```

$$\begin{matrix}
a & b \\
c & d
\end{matrix}$$

### Bracketed Matrices {#sec-bracketed-matrices}

| Environment | Brackets | Example |
|:------------|:---------|:--------|
| `pmatrix` | Parentheses $()$ | $\begin{pmatrix} a & b \\ c & d \end{pmatrix}$ |
| `bmatrix` | Square brackets $[]$ | $\begin{bmatrix} a & b \\ c & d \end{bmatrix}$ |
| `Bmatrix` | Curly braces $\{\}$ | $\begin{Bmatrix} a & b \\ c & d \end{Bmatrix}$ |
| `vmatrix` | Vertical bars (determinant) | $\begin{vmatrix} a & b \\ c & d \end{vmatrix}$ |
| `Vmatrix` | Double vertical bars | $\begin{Vmatrix} a & b \\ c & d \end{Vmatrix}$ |

: Matrix environments {#tbl-matrix-types}

### Example: Tridiagonal Matrix {#sec-example-tridiagonal-matrix}

A more complex matrix with ellipses:

```markdown
$$\begin{matrix}
  -2 & 1 & 0 & 0 & \cdots & 0  \\
  1 & -2 & 1 & 0 & \cdots & 0  \\
  0 & 1 & -2 & 1 & \cdots & 0  \\
  0 & 0 & 1 & -2 & \ddots & \vdots \\
  \vdots & \vdots & \vdots & \ddots & \ddots & 1  \\
  0 & 0 & 0 & \cdots & 1 & -2
\end{matrix}$$
```

$$\begin{matrix}
  -2 & 1 & 0 & 0 & \cdots & 0  \\
  1 & -2 & 1 & 0 & \cdots & 0  \\
  0 & 1 & -2 & 1 & \cdots & 0  \\
  0 & 0 & 1 & -2 & \ddots & \vdots \\
  \vdots & \vdots & \vdots & \ddots & \ddots & 1  \\
  0 & 0 & 0 & \cdots & 1 & -2
\end{matrix}$$

::: {.callout-tip}
In matrix notation:
- Use `&` to separate columns
- Use `\\` to separate rows
- Use `\cdots` for horizontal dots, `\vdots` for vertical, and `\ddots` for diagonal
:::

## Accents and Decorations {#sec-accents}

LaTeX provides various accents and decorations for variables:

| Expression | Code | Description |
|:-----------|:-----|:------------|
| $\hat{x}$ | `$\hat{x}$` | Hat (estimate) |
| $\bar{x}$ | `$\bar{x}$` | Bar (mean) |
| $\tilde{x}$ | `$\tilde{x}$` | Tilde |
| $\vec{x}$ | `$\vec{x}$` | Vector arrow |
| $\dot{x}$ | `$\dot{x}$` | Single dot (derivative) |
| $\ddot{x}$ | `$\ddot{x}$` | Double dot (second derivative) |
| $\mathbf{x}$ | `$\mathbf{x}$` | Bold (vector/matrix) |
| $\mathit{x}$ | `$\mathit{x}$` | Italic |
| $\mathrm{x}$ | `$\mathrm{x}$` | Roman (upright) |

: Mathematical accents and decorations {#tbl-accents}

## Brackets and Delimiters {#sec-brackets}

For equations with tall expressions, use `\left` and `\right` to automatically size delimiters:

```markdown
$$\left( \frac{a^2}{b^2} \right)$$
```

$$\left( \frac{a^2}{b^2} \right)$$

Without automatic sizing:

```markdown
$$( \frac{a^2}{b^2} )$$
```

$$( \frac{a^2}{b^2} )$$

| Delimiter | Code |
|:----------|:-----|
| Parentheses | `\left( ... \right)` |
| Square brackets | `\left[ ... \right]` |
| Curly braces | `\left\{ ... \right\}` |
| Angle brackets | `\left\langle ... \right\rangle` |
| Vertical bars | `\left| ... \right|` |
| Double bars | `\left\| ... \right\|` |

: Bracket delimiters {#tbl-brackets}

## Functions and Operators {#sec-functions}

LaTeX provides proper formatting for common functions:

| Function | Code | Rendered |
|:---------|:-----|:---------|
| $\sin x$ | `$\sin x$` | Sine |
| $\cos x$ | `$\cos x$` | Cosine |
| $\tan x$ | `$\tan x$` | Tangent |
| $\log x$ | `$\log x$` | Logarithm |
| $\ln x$ | `$\ln x$` | Natural log |
| $\exp x$ | `$\exp x$` | Exponential |
| $\lim_{x \to 0}$ | `$\lim_{x \to 0}$` | Limit |
| $\max$ | `$\max$` | Maximum |
| $\min$ | `$\min$` | Minimum |
| $\arg\max$ | `$\arg\max$` | Argument of maximum |

: Mathematical functions {#tbl-functions}

## Numbered and Aligned Equations {#sec-latex-numbered-aligned}

Quarto numbers a display equation when you attach an `{#eq-...}` label after the closing `$$`, and you refer to it with `@eq-...` exactly as for figures and tables (@sec-cross-references):

```markdown
$$
E = mc^2
$$ {#eq-einstein}

@eq-einstein relates mass and energy.
```

$$
E = mc^2
$$ {#eq-einstein-demo}

@eq-einstein-demo relates mass and energy.

A multi-step derivation belongs in an `aligned` environment inside the `$$` pair. Each line ends with `\\`, and the `&` marks the point that lines up vertically (usually just before the `=`):

```markdown
$$
\begin{aligned}
\bar{x} &= \frac{1}{n}\sum_{i=1}^{n} x_i \\
        &= \frac{x_1 + x_2 + \cdots + x_n}{n}
\end{aligned}
$$ {#eq-mean-aligned}
```

$$
\begin{aligned}
\bar{x} &= \frac{1}{n}\sum_{i=1}^{n} x_i \\
        &= \frac{x_1 + x_2 + \cdots + x_n}{n}
\end{aligned}
$$ {#eq-mean-aligned}

The whole block receives one number (@eq-mean-aligned). Use `cases` in the same way for a piecewise definition:

```markdown
$$
f(x) =
\begin{cases}
  \lambda e^{-\lambda x} & x \geq 0 \\
  0                      & x < 0
\end{cases}
$$
```

$$
f(x) =
\begin{cases}
  \lambda e^{-\lambda x} & x \geq 0 \\
  0                      & x < 0
\end{cases}
$$

## Practical Examples {#sec-practical-examples}

### Linear Regression Model {#sec-linear-regression-model}

$$Y_i = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + \epsilon_i$$

where $\epsilon_i \sim N(0, \sigma^2)$.

### Maximum Likelihood {#sec-maximum-likelihood}

$$\hat{\theta} = \arg\max_\theta \prod_{i=1}^{n} f(x_i | \theta)$$

### Bayes' Theorem {#sec-bayes-theorem}

$$P(A|B) = \frac{P(B|A) \cdot P(A)}{P(B)}$$

### Hardy-Weinberg Equilibrium {#sec-hardy-weinberg-equilibrium}

For a biallelic locus with allele frequencies $p$ and $q = 1-p$:

$$p^2 + 2pq + q^2 = 1$$

## Equations From Bioengineering {#sec-latex-bioengineering}

The equations above are the statistician's staples. Here are a few from the kinds of models bioengineers actually write down, each with the LaTeX that produces it, chosen because together they use most of the notation in this chapter.

Enzyme kinetics is the natural starting point. The Michaelis-Menten rate law needs a fraction, a subscript that is a word rather than a single letter (so it is wrapped in `\text{}` or `\mathrm{}`), and square brackets for concentration:

```latex
$$v = \frac{V_{\max}[S]}{K_{\mathrm{M}} + [S]}$$
```

$$v = \frac{V_{\max}[S]}{K_{\mathrm{M}} + [S]}$$

The Hill equation generalizes it to cooperative binding and adds a variable exponent, which shows why multi-character superscripts must be grouped in braces (`[S]^{n}` rather than `[S]^n` when $n$ is more than one character, and always when it is an expression):

```latex
$$\theta = \frac{[L]^{n}}{K_{d} + [L]^{n}}$$
```

$$\theta = \frac{[L]^{n}}{K_{d} + [L]^{n}}$$

Population and cell growth are described by differential equations, and the logistic model shows the derivative notation `\frac{dN}{dt}` together with a parenthesized factor that grows to fit its contents with `\left(` and `\right)`:

```latex
$$\frac{dN}{dt} = rN\left(1 - \frac{N}{K}\right)$$
```

$$\frac{dN}{dt} = rN\left(1 - \frac{N}{K}\right)$$

Transport problems bring in partial derivatives and the Laplacian. Fick's second law of diffusion uses `\partial` and `\nabla^2`; the second form, written out in one dimension, is the one you would actually discretize in code:

```latex
$$\frac{\partial C}{\partial t} = D\,\nabla^{2} C
  \qquad\text{or, in one dimension,}\qquad
  \frac{\partial C}{\partial t} = D\frac{\partial^{2} C}{\partial x^{2}}$$
```

$$\frac{\partial C}{\partial t} = D\,\nabla^{2} C
  \qquad\text{or, in one dimension,}\qquad
  \frac{\partial C}{\partial t} = D\frac{\partial^{2} C}{\partial x^{2}}$$

Finally, the normalization used on the course dataset in @sec-writing-functions is itself an equation worth writing down, because doing so makes the pseudocount explicit in a way the code never quite does. For gene $g$ in sample $i$, with raw count $c_{gi}$ and library size $N_i = \sum_g c_{gi}$,

```latex
$$\text{CPM}_{gi} = \frac{c_{gi}}{N_i}\times 10^{6},
  \qquad
  y_{gi} = \log_{2}\!\left(\text{CPM}_{gi} + 1\right)$$
```

$$\text{CPM}_{gi} = \frac{c_{gi}}{N_i}\times 10^{6},
  \qquad
  y_{gi} = \log_{2}\!\left(\text{CPM}_{gi} + 1\right)$$

A methods section that carries this equation, with the code that implements it in the same Quarto document, is the kind of reproducibility the last chapters of this book are about.

## Reading Equations You Did Not Write {#sec-latex-reading}

You will spend more time reading LaTeX than writing it: in the methods sections of papers whose source you download, in the templates of the journal you submit to, and in code comments where an author has written the formula their function implements. Three habits make that reading easy. First, find the structure before the symbols: every `\frac{...}{...}` is one fraction with a numerator and a denominator, every `\sum_{...}^{...}` is one sum with a lower and an upper limit, and matching braces tell you where each piece begins and ends, so an expression that looks like a wall of backslashes is usually three or four nested pieces. Second, learn to see a dozen commands on sight, since they account for most of what appears in scientific writing: the Greek letters, `\frac`, `\sqrt`, `\sum`, `\int`, `\partial`, `\hat`, `\bar`, `\left` and `\right`, `\text`, and the subscript and superscript operators `_` and `^`. Third, when a symbol is unfamiliar, render it. Paste the expression between dollar signs in a Quarto document and preview it, or use one of the online detexify tools that let you draw a symbol and tell you its command. Appendix @sec-appendix-latex is organized for exactly this kind of look-up.

::: {.callout-note title="LaTeX beyond Quarto"}
Quarto lets you use LaTeX's mathematics without writing a LaTeX document. You will meet the full system eventually: most journals in engineering and the quantitative sciences supply a LaTeX template, theses at many universities are written in it, and **Overleaf** (<https://www.overleaf.com/learn>) is a browser-based editor that many collaborations use so that nobody has to install anything. The mathematics you learn here transfers unchanged; what a full LaTeX document adds is the document-level structure of `\documentclass`, `\section`, and `\begin{figure}` that Quarto has been writing for you behind the scenes whenever you render to PDF (@sec-quarto-documents).
:::

## Best Practices {#sec-latex-best-practices}

### When to Use LaTeX {#sec-latex-when-to-use-latex}

- Use inline math (`$...$`) for simple expressions in text
- Use display math (`$$...$$`) for important formulas or complex expressions
- Use display math when equations would break text flow

### Formatting Tips {#sec-formatting-tips}

1. **Spacing**: LaTeX ignores spaces in math mode. Use `\,` for a thin space (as before $dx$ in an integral), `\:` for medium, `\;` for thick, `\quad` and `\qquad` for large gaps, and `\!` to pull symbols closer (full table in @sec-latex-spacing)
2. **Text in equations**: Use `\text{...}` for words: $P(\text{event})$
3. **Size**: Use `\large`, `\small`, etc. for size adjustments
4. **Numbering**: In Quarto, add `{#eq-label}` after display equations for cross-referencing (@sec-latex-numbered-aligned)

### Common Mistakes {#sec-common-mistakes}

::: {.callout-warning}
**Avoid these common errors:**

- Forgetting to escape braces: `\{` not `{` when you want literal braces
- Missing closing delimiters with `\left` and `\right`
- Using `*` for multiplication instead of `\times` or `\cdot`
- Not grouping multi-character subscripts/superscripts with braces
:::

## Summary {#sec-latex-summary}

LaTeX provides powerful tools for mathematical typesetting (a complete symbol reference is in @sec-appendix-latex):

- **Inline** (`$...$`) and **display** (`$$...$$`) modes for different contexts
- **Greek letters** and **special symbols** with backslash commands
- **Fractions**, **roots**, and **powers** for complex expressions
- **Sums**, **products**, and **integrals** with limits
- **Matrices** with various bracket styles
- **Automatic sizing** of delimiters with `\left` and `\right`
- **Numbered** (`{#eq-...}`) and **aligned** (`aligned`, `cases`) display equations

Integration with Quarto and R Markdown means you can include precise mathematical notation alongside code and narrative—essential for reproducible scientific communication.

## Exercises {#sec-latex-exercises .unnumbered}

::: {.callout-tip title="Practice Exercises"}
**Exercise 1: Basic Notation**

Write the LaTeX code to produce these expressions:

1. $x^2 + y^2 = z^2$
2. $\frac{-b \pm \sqrt{b^2-4ac}}{2a}$
3. $\sum_{i=1}^{n} x_i^2$

**Exercise 2: Statistical Formulas**

Write the LaTeX for these statistical expressions:

1. Sample correlation coefficient
2. Chi-square test statistic
3. A 95% confidence interval

**Exercise 3: Matrix Operations**

Create a $3 \times 3$ identity matrix using `bmatrix` environment.

**Exercise 4: Integrate into Quarto**

Create a Quarto document that:

1. Explains a statistical concept with inline equations
2. Includes a centered display equation for the main formula
3. Uses appropriate Greek letters and symbols
:::
