---
title: "Computational Lab Notebook"
subtitle: "Project: [short project name]"
author: "Your Name"
date: today
format:
  html:
    toc: true
    embed-resources: true
    code-tools: true
    df-print: paged
execute:
  warning: false
  message: false
---

<!--
HOW TO USE THIS TEMPLATE
========================
A computational lab notebook plays the same role as a wet-lab
notebook: a dated, honest record of what you did, why you did it,
what happened, and what you concluded — including the dead ends.

Suggested workflow:
  1. Copy this file and rename it, e.g. 2026-10-06_rnaseq-qc.qmd
     (one notebook file per experiment or analysis session).
  2. Fill in each section as you work, not after the fact.
  3. Render to HTML (Cmd/Ctrl + Shift + K in Positron) when done;
     with embed-resources: true the HTML is a single, self-contained
     file you can archive or share.
  4. Commit both the .qmd and rendered .html to version control.

Everything below the "worked example" line is a filled-in entry
using the course stickleback dataset — replace it with your own.
-->

# Entry: `r format(Sys.Date())` — [one-line title of this analysis]

## 1. Objective

*What question is this entry trying to answer? One or two sentences.
If you can't state the objective, you aren't ready to write code.*

## 2. Background and rationale

*Why are you doing this now? Link to prior entries, papers, or lab
discussions. What did you already know before starting?*

## 3. Data provenance

*Exactly where the data came from — future-you will not remember.*

| Item | Value |
|------|-------|
| File(s) | `path/to/data.tsv` |
| Source | e.g. sequencing core, collaborator, NCBI accession |
| Downloaded / received | YYYY-MM-DD |
| Version / md5 checksum | `md5sum path/to/data.tsv` |

## 4. Methods and parameters

*Software, versions, and every non-default parameter. The code
chunks below are part of the record, but summarize key choices in
prose so a reader doesn't have to reverse-engineer them.*

## 5. Analysis

*Interleave narrative and code. Say what each chunk is about to do
and, after it runs, what you see in the output.*

```{r}
#| label: setup
library(readr)
library(dplyr)
library(ggplot2)
```

## 6. Results and interpretation

*What did you find? Refer to specific numbers and figures. Separate
observation ("the interaction term is significant, p = 3e-7") from
interpretation ("the microbiota effect depends on host genotype").*

## 7. Problems and dead ends

*Things that failed, surprised you, or were abandoned — and why.
This is the section that makes a notebook honest and useful.*

## 8. Next steps

*Concrete, actionable items for the next session.*

## 9. Reproducibility record

```{r}
#| label: session-info
sessionInfo()
```

---

# Worked example entry: 2026-09-29 — Genotype and microbiota effects on Gene369

## 1. Objective

Test whether gut expression of Gene369 depends on host genotype,
microbiota treatment, and their interaction, in the course RNA-seq
dataset of 80 stickleback.

## 2. Background and rationale

Gene369 showed high between-group variance in an initial screen of
the 600-gene count table. The experiment crossed 2 host genotypes
(A, B) with 2 microbiota treatments (conventional, mono-associated),
n = 20 fish per group, so a two-factor linear model is the natural
first analysis.

## 3. Data provenance

| Item | Value |
|------|-------|
| File | `Data_Folder/Stickle_RNAseq.tsv` |
| Source | Course website (Cresko Lab gut RNA-seq counts) |
| Downloaded | 2026-09-29 |
| Shape | 80 fish x 604 columns (3 factors + 600 genes) |

## 4. Methods and parameters

Counts were normalized to log2 counts-per-million (CPM) with a
pseudocount of 1, using each fish's total count as its library
size. Effects were tested with an ordinary least-squares linear
model, `gene369 ~ Genotype * Microbiota`, in base R.

## 5. Analysis

Read the counts and normalize the focal gene:

```{r}
#| label: read-normalize
local_paths <- c(
  "Data_Folder/Stickle_RNAseq.tsv",
  "../Data_Folder/Stickle_RNAseq.tsv"
)
found <- local_paths[file.exists(local_paths)]
data_file <- if (length(found) > 0) {
  found[1]
} else {
  "https://wcresko.github.io/BioE_Comp_Tools_Fa2026/Data_Folder/Stickle_RNAseq.tsv"
}

stickle <- read_tsv(data_file, show_col_types = FALSE) |>
  mutate(lib_size = rowSums(across(starts_with("Gene"))))

focal <- stickle |>
  transmute(
    Individual, Genotype, Microbiota,
    gene369 = log2(Gene369 / lib_size * 1e6 + 1)
  )

count(focal, Genotype, Microbiota)
```

Visualize expression by group before modeling:

```{r}
#| label: fig-boxplot
#| fig-cap: "Gene369 expression (log2 CPM) by genotype and microbiota treatment."
ggplot(focal, aes(x = Microbiota, y = gene369, fill = Genotype)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_point(
    position = position_jitterdodge(jitter.width = 0.15),
    size = 1.5, alpha = 0.6
  ) +
  labs(y = "Gene369 expression (log2 CPM)") +
  theme_minimal()
```

Fit the two-factor model with interaction:

```{r}
#| label: linear-model
fit <- lm(gene369 ~ Genotype * Microbiota, data = focal)
summary(fit)
anova(fit)
```

Check the residuals:

```{r}
#| label: fig-residuals
#| fig-cap: "Residuals vs. fitted values for the two-factor model."
plot(fit, which = 1)
```

## 6. Results and interpretation

Both main effects and the interaction are significant. Genotype B
fish express Gene369 about 4.5 log2 CPM units lower than genotype A
(p < 1e-30), mono-association lowers expression by about 2.8 units
in genotype A (p < 1e-20), and the positive interaction term
(+1.5, p = 3e-7) means the microbiota effect is weaker in genotype
B. The model explains ~93% of the variance. Interpretation: the
transcriptional response of Gene369 to microbiota depends on host
genotype — a candidate for genotype-by-environment interaction.

## 7. Problems and dead ends

Raw counts were initially modeled without normalization; library
sizes vary ~4.6-fold across fish, which inflated the residual
variance. Switched to log2 CPM.

## 8. Next steps

- Repeat the model across all 600 genes; adjust p-values (FDR).
- Check whether Gene369's co-expression partners (e.g. Gene208)
  show the same interaction pattern.

## 9. Reproducibility record

```{r}
#| label: session-info-example
sessionInfo()
```
