---
title: "Lecture 06 - Computational Tools for Bioengineers"
subtitle: "R and Python Fundamentals 2: Data Frames, Distributions and Figures"
date: "Monday, November 2, 2026"
author: "Bill Cresko"
---

```{r}
#| label: setup
#| include: false

# Packages used in the lecture examples
library(tidyverse)
library(gt)
library(readxl)

# Consistent theme for all plots
theme_set(theme_minimal(base_size = 14))

# Reproducible random numbers
set.seed(2026)
```



## Book reading for this lecture

The course book goes deeper on everything in this lecture. Read before or after class:

-   [Ch. 11: R Programming Fundamentals](../book/chapters/11-r-programming.html)
-   [Ch. 12: Python Programming Fundamentals](../book/chapters/12-python-programming.html)
-   [Ch. 13: Data Visualization with ggplot2](../book/chapters/13-data-visualization.html)
-   *Reference:* [App. E: R Command Reference](../book/chapters/appendix-E-r.html) · [App. F: Python Quick Reference](../book/chapters/appendix-F-python.html)

## Goals for today

-   Reading, creating and exporting data frames
-   Drawing samples from distributions
-   Visualizing data and combining plots into one figure
-   Indexing data frames and computing summary statistics

## Reading in Data Frames in R

-   A strength of `R` is being able to import data from an external source
-   Create the same table that you did above in a spreadsheet like Excel
-   Export it to comma separated and tab separated text files for importing into `R`.
-   The first will read in a comma-delimited file, whereas the second is a tab-delimited
-   In both cases the header and row.names arguments indicate that there is a header row and row label column
-   Note that the name of the file by itself will have R look in the CWD, whereas a full path can also be used
-   In Python the same job is done by **pandas**: `pd.read_csv()` for comma-separated, `pd.read_csv(sep="\t")` for tab-separated; `pathlib.Path` builds paths that work on every operating system

## Drawing samples from distributions

-   Here is a way to create your own data sets that are random samples.
-   Again, play around with the arguments in the parentheses to see what happens.
-   In Python random numbers come from a numpy **generator**: `rng = np.random.default_rng(2026)` (the number is the seed, like `set.seed(2026)` in R)

## Drawing samples from distributions

::: panel-tabset
### R

```{r}
#| label: fig-lec07-samples-from-distributions-1
#| fig-cap: "10,000 normal draws (x) plotted against 10,000 uniform draws (y)"
#| out-width: "100%"
#| echo: true
#| eval: true

x <- rnorm (10000, 0, 10)
y <- sample (1:10000, 10000, replace = TRUE)
xy <- cbind(x,y)
plot(x,y)
```

### Python

```{python}
#| label: fig-lec07-samples-from-distributions-1-py
#| fig-cap: "10,000 normal draws (x) plotted against 10,000 uniform draws (y), with matplotlib"
#| echo: true
#| eval: false

import numpy as np
import matplotlib.pyplot as plt

rng = np.random.default_rng(2026)
x = rng.normal(0, 10, 10000)               # rnorm(10000, 0, 10)
y = rng.integers(1, 10001, size=10000)     # sample(1:10000, 10000, replace = TRUE)
xy = np.column_stack([x, y])               # cbind(x, y)
plt.scatter(x, y, s=2)
plt.show()
```
:::

## Drawing samples from distributions

::: panel-tabset
### R

```{r}
#| label: fig-lec07-samples-from-distributions-2
#| fig-cap: "The same samples plotted from a two-column matrix made with cbind()"
#| out-width: "100%"
#| echo: true
#| eval: true

x <- rnorm (10000, 0, 10)
y <- sample (1:10000, 10000, replace = TRUE)
xy <- cbind(x,y)
plot(xy)
```

### Python

```{python}
#| label: fig-lec07-samples-from-distributions-2-py
#| fig-cap: "The same samples plotted from the columns of a two-column array"
#| echo: true
#| eval: false

xy = np.column_stack([x, y])
print(xy.shape)
# (10000, 2)
plt.scatter(xy[:, 0], xy[:, 1], s=2)       # column 0 vs column 1
plt.show()
```
:::

## Drawing samples from distributions

::: panel-tabset
### R

```{r}
#| label: fig-lec07-samples-from-distributions-3
#| fig-cap: "Histogram of 10,000 draws from a normal distribution (mean 0, sd 10)"
#| out-width: "100%"
#| echo: true
#| eval: true

x <- rnorm (10000, 0, 10)
y <- sample (1:10000, 10000, replace = TRUE)
xy <- cbind(x,y)
hist(x)
```

### Python

```{python}
#| label: fig-lec07-samples-from-distributions-3-py
#| fig-cap: "Histogram of 10,000 draws from a normal distribution (mean 0, sd 10), with matplotlib"
#| echo: true
#| eval: false

plt.hist(x, bins=30)
plt.xlabel("x")
plt.ylabel("Frequency")
plt.show()
```
:::

## Drawing samples from distributions

-   You’ve probably figured out that y from the last example is drawing numbers with equal probability.
-   What if you want to draw from a distribution?
-   Again, play around with the arguments in the parentheses to see what happens.

## Drawing samples from distributions

::: panel-tabset
### R

```{r}
#| label: fig-lec07-drawing-samples-from-distributions
#| fig-cap: "Histogram of normal draws with the theoretical density curve overlaid"
#| out-width: "100%"
#| echo: true
#| eval: true

x <-rnorm(1000, 0, 100)
hist(x, xlim = c(-500,500))
curve(50000*dnorm(x, 0, 100), xlim = c(-500,500), add=TRUE, col='Red')
```

### Python

```{python}
#| label: fig-lec07-drawing-samples-from-distributions-py
#| fig-cap: "Histogram of normal draws with the theoretical density curve overlaid (scipy.stats)"
#| echo: true
#| eval: false

from scipy.stats import norm

x = rng.normal(0, 100, 1000)
plt.hist(x, bins=20, range=(-500, 500))    # bins 50 wide, like R's default here
grid = np.linspace(-500, 500, 200)
plt.plot(grid, 50000 * norm.pdf(grid, 0, 100), color="red")
plt.show()
```
:::

-   `dnorm()` generates the probability density, which can be plotted using the `curve()` function.
-   Note that is curve is added to the plot using `add=TRUE`
-   In Python `norm.pdf()` from `scipy.stats` is the density; every `plt.` call draws on the same figure until `plt.show()`

## Visualizing Data

-   So far you've been visualizing just the list of output numbers
-   Except for the last example where I snuck in a `hist` function.
-   You can also visualize all of the variables that you've created using the `plot` function (as well as a number of more sophisticated plotting functions).
-   Each of these is called a `high level` plotting function, which sets the stage
-   `Low level` plotting functions will tweak the plots and make them beautiful

## Visualizing Data

-   What do you think that each of the arguments means for the plot function?
-   A cool thing about `R` is that the options for the arguments make sense.
-   Try adjusting an argument and see if it works
-   Note- most people use the package `GGPlot2` for nice scientific graphics
-   In Python the base library is `matplotlib` (`plt.plot`, `plt.scatter`, `plt.hist`); `seaborn` and `plotnine` (a ggplot2 clone) sit on top of it

## Visualizing Data

::: panel-tabset
### R

```{r}
#| label: fig-lec07-visualizing-data
#| fig-cap: "A sequence from 0 to 10 plotted as points"
#| echo: true
#| eval: true

seq_1 <- seq(0.0, 10.0, by = 0.1)
plot (seq_1, xlab="space", ylab ="function of space", type = "p", col = "red")
```

### Python

```{python}
#| label: fig-lec07-visualizing-data-py
#| fig-cap: "A sequence from 0 to 10 plotted as points, with matplotlib"
#| echo: true
#| eval: false

seq_1 = np.linspace(0.0, 10.0, 101)
plt.plot(seq_1, "o", color="red")          # "o" = points; default is a line
plt.xlabel("space")
plt.ylabel("function of space")
plt.show()
```
:::

## Putting plots in a single figure

-   On the next slide
-   The first line of the lower script tells R that you are going to create a composite figure that has two rows and two columns. Can you tell how?
-   In Python, `plt.subplots(2, 2)` does the same job and hands back a grid of **axes** to draw on
-   Now, modify the code to add two more variables and add one more row of two panels.

::: panel-tabset
### R

```{r}
#| label: lec07-putting-plots-in-a-single
#| out-width: "50%"
#| out-height: "20%"
#| echo: true
#| eval: true

seq_1 <- seq(0.0, 10.0, by = 0.1)
seq_2 <- seq(10.0, 0.0, by = -0.1)
```

### Python

```{python}
#| label: lec07-putting-plots-in-a-single-py
#| echo: true
#| eval: false

seq_1 = np.linspace(0.0, 10.0, 101)
seq_2 = np.linspace(10.0, 0.0, 101)
seq_square = seq_2 * seq_2
seq_square_new = seq_2**2
```
:::

## Putting plots in a single figure

::: panel-tabset
### R

```{r}
#| label: fig-lec07-composite-figure
#| fig-cap: "Four plots arranged in a 2 x 2 grid with par(mfrow)"
#| out-width: "50%"
#| out-height: "50%"
#| echo: true
#| eval: true

par(mfrow=c(2,2))
plot (seq_1, xlab="time", ylab ="p in population 1", type = "p", col = 'red')
plot (seq_2, xlab="time", ylab ="p in population 2", type = "p", col = 'green')
plot (seq_square, xlab="time", ylab ="p2 in population 2", type = "p", col = 'blue')
plot (seq_square_new, xlab="time", ylab ="p in population 1", type = "l", col = 'yellow')
```

### Python

```{python}
#| label: fig-lec07-composite-figure-py
#| fig-cap: "Four plots arranged in a 2 x 2 grid with plt.subplots()"
#| echo: true
#| eval: false

fig, ax = plt.subplots(2, 2, figsize=(8, 6))
ax[0, 0].plot(seq_1, "o", color="red")
ax[0, 0].set(xlabel="time", ylabel="p in population 1")
ax[0, 1].plot(seq_2, "o", color="green")
ax[0, 1].set(xlabel="time", ylabel="p in population 2")
ax[1, 0].plot(seq_square, "o", color="blue")
ax[1, 0].set(xlabel="time", ylabel="p2 in population 2")
ax[1, 1].plot(seq_square_new, "-", color="yellow")   # "-" = line
ax[1, 1].set(xlabel="time", ylabel="p in population 1")
fig.tight_layout()
plt.show()
```
:::

## Creating Data Frames in R and Python

-   As you have seen, in R you can generate your own random data set drawn from nearly any distribution very easily.
-   Often we will want to use collected data.
-   Now, let’s make a dummy dataset to get used to dealing with data frames
-   Set up three variables (hydrogel_concentration, compression and conductivity) as vectors
-   In Python the data frame comes from **pandas** (`import pandas as pd`); it has the same idea - rows are samples, columns are variables

## Creating Data Frames in R and Python

::: panel-tabset
### R

```{r}
#| label: lec07-creating-data-frames-in-r
#| echo: true
#| eval: true

hydrogel_concentration <- factor(c("low", "high", "high", "high", "medium", "medium", "medium","low"))
compression <- c(3.4, 3.4, 8.4, 3, 5.6, 8.1, 8.3, 4.5)
conductivity <- c(0, 9.2, 3.8, 5, 5.6, 4.1, 7.1, 5.3)
```

-   Create a data frame where vectors become columns

```{r}
#| label: lec07-creating-data-frames-in-r-2
#| echo: true
#| eval: true

mydata <- data.frame(hydrogel_concentration, compression, conductivity)
row.names(mydata) <- c("Sample_1", "Sample_2", "Sample_3", "Sample_4",
                       "Sample_5", "Sample_6", "Sample_7", "Sample_8")
```

### Python

```{python}
#| label: lec07-creating-data-frames-py
#| echo: true
#| eval: false

import pandas as pd

hydrogel_concentration = pd.Categorical(["low", "high", "high", "high", "medium", "medium", "medium", "low"])
compression = [3.4, 3.4, 8.4, 3, 5.6, 8.1, 8.3, 4.5]
conductivity = [0, 9.2, 3.8, 5, 5.6, 4.1, 7.1, 5.3]
```

-   Create a DataFrame from a dictionary of `column name: values`

```{python}
#| label: lec07-creating-data-frames-2-py
#| echo: true
#| eval: false

mydata = pd.DataFrame({"hydrogel_concentration": hydrogel_concentration,
                       "compression": compression,
                       "conductivity": conductivity})
mydata.index = [f"Sample_{i}" for i in range(1, 9)]   # row names = the index
mydata.head(3)
#          hydrogel_concentration  compression  conductivity
# Sample_1                    low          3.4           0.0
# Sample_2                   high          3.4           9.2
# Sample_3                   high          8.4           3.8
```
:::

-   Now you have a hand-made data frame with row names
-   Take a look at it in the data section of RStudio, or the **Data Explorer** in Positron (R and Python)

## Reading in and Exporting Data Frames

\

::: panel-tabset
### R

```{r}
#| label: lec07-reading-in-and-exporting-data
#| echo: true
#| eval: false

YourFile <- read.table('yourfile.csv', header=T, row.names=1, sep=',')
YourFile <- read.table('yourfile.txt', header=T, row.names=1, sep='\t')
```

\

```{r}
#| label: lec07-reading-in-and-exporting-data-2
#| echo: true
#| eval: false

write.table(YourFile, "yourfile.csv", quote=F, row.names=T, sep=",")
write.table(YourFile, "yourfile.txt", quote=F, row.names=T, sep="\t")
```

### Python

```{python}
#| label: lec07-reading-in-and-exporting-data-py
#| echo: true
#| eval: false

import pandas as pd
from pathlib import Path

YourFile = pd.read_csv("yourfile.csv", index_col=0)             # header row is the default
YourFile = pd.read_csv("yourfile.txt", index_col=0, sep="\t")
YourFile = pd.read_csv(Path("data") / "raw" / "yourfile.csv")   # portable paths with pathlib
```

\

```{python}
#| label: lec07-reading-in-and-exporting-data-2-py
#| echo: true
#| eval: false

YourFile.to_csv("yourfile.csv")                 # index (row names) is written by default
YourFile.to_csv("yourfile.txt", sep="\t")
YourFile.to_csv("yourfile.csv", index=False)    # drop the row names
```
:::

## Indexing in data frames

-   Next up - indexing just a subset of the data
-   This is a very important idea in R, that you can analyze just a subset of the data.
-   This is analyzing only the data in the file you made that has the factor value 'mixed'.
-   pandas: `.iloc[row, col]` indexes by **position** (from 0), `.loc[row, col]` by **name**; `df["col"]` or `df.col` picks a column

::: panel-tabset
### R

```{r}
#| label: lec07-indexing-in-data-frames
#| echo: true
#| eval: false

print (YourFile[,2])
print (YourFile$variable)
print (YourFile[2,])
plot (YourFile$variable1, YourFile$variable2)
```

### Python

```{python}
#| label: lec07-indexing-in-data-frames-py
#| echo: true
#| eval: false

print(YourFile.iloc[:, 1])          # 2nd column (position 1) - R's YourFile[,2]
print(YourFile["variable"])         # column by name        - R's YourFile$variable
print(YourFile.iloc[1])             # 2nd row               - R's YourFile[2,]
print(YourFile.loc["Sample_2"])     # row by its name
plt.scatter(YourFile["variable1"], YourFile["variable2"])
plt.show()
```
:::

## Practice reading in transcriptomic data

::: panel-tabset
### R

```{r}
#| label: lec07-practice-reading-in-transcriptomic-data
#| echo: true
#| eval: false

RNAseq_Data <- read.table('<name_of_file>', header=TRUE, sep=',')

print (RNAseq_Data)
head (RNAseq_Data)
tail (RNAseq_Data)

print (RNAseq_Data[,2])
print (RNAseq_Data[1,])
print (RNAseq_Data[1,2])
print (RNAseq_Data$ENSGACG00000000010)
print (RNAseq_Data$ENSGACG00000000010>45.0)
```

### Python

```{python}
#| label: lec07-practice-reading-in-transcriptomic-data-py
#| echo: true
#| eval: false

RNAseq_Data = pd.read_csv("<name_of_file>")

print(RNAseq_Data)
RNAseq_Data.head()
RNAseq_Data.tail()

print(RNAseq_Data.iloc[:, 1])
print(RNAseq_Data.iloc[0])          # first row is 0!
print(RNAseq_Data.iloc[0, 1])
print(RNAseq_Data["ENSGACG00000000010"])
print(RNAseq_Data["ENSGACG00000000010"] > 45.0)
```
:::

## Summary stats and figures

::: panel-tabset
### R

```{r}
#| label: lec07-summary-stats-and-figures
#| echo: true
#| eval: false

summary1 <- summary(RNAseq_Data $ENSGACG00000000003)
print (summary1)

hist(RNAseq_Data $ENSGACG00000000003)
boxplot(RNAseq_Data$ENSGACG00000000003)
boxplot(RNAseq_Data$ENSGACG00000000003~RNAseq_Data$Population)
plot(RNAseq_Data $ENSGACG00000000003, RNAseq_Data$ENSGACG00000000003)

boxplot(RNAseq_Data $ENSGACG00000000003~RNAseq_Data$Treatment,
        col = "red", ylab = "Expression Level", xlab = "Treatment level",
        border ="orange",
        main = "Boxplot of variation in gene expression across microbiota treatments")
```

### Python

```{python}
#| label: lec07-summary-stats-and-figures-py
#| echo: true
#| eval: false

import seaborn as sns

gene = RNAseq_Data["ENSGACG00000000003"]
summary1 = gene.describe()          # count, mean, std, min, quartiles, max
print(summary1)

gene.hist()
gene.plot.box()
RNAseq_Data.boxplot(column="ENSGACG00000000003", by="Population")
plt.scatter(gene, gene)

sns.boxplot(data=RNAseq_Data, x="Treatment", y="ENSGACG00000000003", color="red")
plt.xlabel("Treatment level")
plt.ylabel("Expression Level")
plt.title("Boxplot of variation in gene expression across microbiota treatments")
plt.show()
```
:::
