---
title: "Lecture 07 - Computational Tools for Bioengineers"
subtitle: "R and RStudio"
date: "Monday, November 9, 2026"
author: "Bill Cresko"
---

```{r}
#| label: setup
#| include: false

# Packages used in the lecture examples
library(tidyverse)
library(gt)
library(readxl)

# Consistent theme for all plots
theme_set(theme_minimal(base_size = 14))

# Reproducible random numbers
set.seed(2026)
```


## Goals for today

-   Using R Studio and Visual Studio Code (VSC) as script editors
-   R and Quarto Markdown
-   Vibe Coding with LLMs
-   Git and GitHub

## What is a Shell Script?

-   Text file containing Unix commands
-   Automates repetitive tasks
-   Makes complex workflows reproducible
-   Essential for bioinformatics pipelines
-   Can include:
    -   Variables and arrays
    -   Conditionals (if/then/else)
    -   Loops (for/while)
    -   Functions
    -   Command-line arguments

## Creating Your First Script

``` bash
#!/bin/bash
# This is a shebang - tells system which interpreter to use

# This is a comment - documents your code
echo "Hello, Bioengineering!"

# Set a variable
NAME="DNA Analysis"
echo "Running $NAME"

# Use command output as variable
DATE=$(date +%Y-%m-%d)
echo "Analysis date: $DATE"

# Simple calculation
COUNT=5
DOUBLE=$((COUNT * 2))
echo "Double of $COUNT is $DOUBLE"
```

Save as `first_script.sh` and run with `bash first_script.sh`

## Making Scripts Executable

``` bash
# Check current permissions
ls -l first_script.sh

# Make executable (add execute permission)
chmod +x first_script.sh

# Now you can run it directly
./first_script.sh
```

Best practice: Always use `chmod +x` for your scripts

# R statistical programming language

## Why use `R`?

-   Good general scripting tool for statistics and mathematics
-   Powerful and flexible and free
-   Runs on all computer platforms
-   New enhancements coming out all the time
-   Superb data management & graphics capabilities
-   Reproducibility - can keep your scripts to see exactly what was done
-   You can write your own functions
-   Lots of online help available
-   Can use a nice GUI front end such as `Rstudio`
-   Can embed your `R` analyses in dynamic, polished files using `Quarto Markdown`
-   Quarto Markdown can be reused for websites, papers, books, presentations...

## `R scripts` and `Quarto Markdown`

-   Often we want to write scripts that can just be run - `R scripts`
-   We can also embed code in Quarto Markdown files that provide more annotations
-   This improves interpertability and reproducibility
-   You can insert `R code chunks` into `Quarto Markdown` documents

## `R script` basics

-   A series of R commands that will be executed\

-   Can add comments using hashtags `#`\

-   Can have pipes (`|>`) to connect one step to the next\

## BASICS of `R`

-   Commands can be submitted through the terminal, console or scripts
-   Notice on these slides I'm evaluating the code chunks and showing output
-   Also notice that `R` follows the normal priority of mathematical evaluation

\
\

```{r}
#| label: lec07-basic-multiplication
#| echo: true
#| eval: true

4*4
```

```{r}
#| label: lec07-more-multiplication
#| echo: true
#| eval: true

(4+3*2^2)
```

## Assigning Variables

-   A better way to do this is to assign variables
-   Variables are assigned values using the `<-` operator.
-   Variable names must begin with a letter, but other than that, just about anything goes.
-   Do keep in mind that `R` is case sensitive.

## Assigning Variables

```{r}
#| label: lec07-assigning-variables
#| echo: true
#| eval: true

x <- 2
x*3
y <- x * 3
y-2

```

These do not work

```{r}
#| label: lec07-assigning-variables-2
#| echo: true
#| eval: false

3y <- 3
3*y <- 3
```

## Arithmetic operations on functions

-   Arithmetic operations can be performed easily on functions as well as numbers.
-   Try the following, and then your own.

```{r}
#| label: lec07-functions
#| echo: true
#| eval: false

x+2
x^2
log(x)
```

-   Note that the last of these - `log` - is a built in function of `R`, and therefore the object of the function needs to be put in parentheses
-   These parentheses will be important, and we'll come back to them later when we add arguments after the object in the parentheses\
-   The outcome of calculations can be assigned to new variables as well, and the results can be checked using the 'print' command

## Arithmetic operations on functions

```{r}
#| label: lec07-arithmetic-operations-on-functions
#| echo: true
#| eval: true

y <- 67
print(y)

x <- 124
z <- (x*y)^2
print(z)
```

## STRINGS

-   Variables and operations can be performed on characters as well
-   Note that characters need to be set off by quotation marks to differentiate them from numbers
-   The `c` stands for `concatenate`
-   Note that we are using the same variable names as we did previously, which means that we're overwriting our previous assignment
-   A good rule of thumb is to use new names for each variable, and make them short but still descriptive

## STRINGS

```{r}
#| label: lec07-strings
#| echo: true
#| eval: true

x <- "I Love"
print (x)
y <- "Bioengineering"
print (y)
z <- c(x,y)
print (z)
```

## VECTORS

-   In general `R` thinks in terms of vectors (a list of characters, factors or numerical values).
-   It will benefit any `R` user to try to write programs with that in mind, as it will simplify most things.
-   Vectors can be assigned directly using the `c()` function and then entering the exact values.

## VECTORS

```{r}
#| label: lec07-example-vectors
#| echo: true
#| eval: true

x <- c(2,3,4,2,1,2,4,5,10,8,9)
print(x)
```

```{r}
#| label: lec07-example-vector-multiplication
#| echo: true
#| eval: true

x <- c(2,3,4,2,1,2,4,5,10,8,9)
y <- x^2
print(x)
print(y)
```

```{r}
#| label: fig-lec07-example-vector-plot
#| fig-cap: "Each value of x plotted against its square (y = x^2)"
#| echo: true
#| eval: true

plot(x,y)
```

## Basic Statistics

-   Many functions exist to operate on vectors.
-   Combine these with your previous variable to see what happens.
-   Also, try to find other functions (e.g. standard deviation).

```{r}
#| label: lec07-basic-statistics
#| echo: true
#| eval: false

mean(x)
median(x)
var(x)
log(x)
ln(x)
sqrt(x)
sum(x)
length(x)
sample(x, replace = T)
```

-   Notice that the last function (`sample`) has an argument (`replace=T`)
-   Arguments simply modify or direct the function in some way
-   There are many arguments for each function, some of which are defaults

## Getting Help

-   Getting Help on any function is very easy - just type a question mark and the name of the function.
-   There are functions for just about anything within `R` and it is easy enough to write your own functions if none already exist to do what you want to do.
-   In general, function calls have a simple structure: a function name, a set of parentheses and an optional set of parameters to send to the function.
-   Help pages exist for all functions that, at a minimum, explain what parameters exist for the function.\
-   Help can be accessed a few ways - try them :

## Getting Help

```{r}
#| label: lec07-getting-help
#| echo: true
#| eval: false

- help(mean)
- ?mean
- example(mean)
- help.search("mean")
- apropos("mean")
- args(mean)
```

## Creating vectors

-   Creating vector of new data by entering it by hand can be a drag
-   However, it is also very easy to use functions such as `seq` and `sample`
-   Try the examples below; can you figure out what the three arguments in the parentheses mean?
-   Try varying the arguments to see what happens.
-   Don't go too crazy with the last one or your computer might slow way down

## Creating vectors

```{r}
#| label: lec07-creating-vectors
#| echo: true
#| eval: true

seq_1 <- seq(0.0, 10.0, by = 0.1)
print(seq_1)
seq_2 <- seq(10.0, 0.0, by = -0.1)
print(seq_2)
```

## Creating vectors

```{r}
#| label: lec07-creating-vectors-2
#| echo: true
#| eval: true

seq_square <- (seq_2)*(seq_2)
print(seq_square)
```

## Creating vectors

```{r}
#| label: lec07-creating-vectors-3
#| echo: true
#| eval: true

seq_square_new <- (seq_2)^2
print(seq_square_new)
```

## Types of vectors of data

-   `int` stands for integers

-   `dbl` stands for doubles, or real numbers

-   `chr` stands for character vectors, or strings

-   `dttm` stands for date-times (a date + a time)

-   `lgl` stands for logical, vectors that contain only TRUE or FALSE

-   `fctr` stands for factors, which R uses to represent categorical variables with fixed possible values

-   `date` stands for dates

<!-- -->

-   Integer and double vectors are known collectively as numeric vectors.
-   In `R` numbers are doubles by default.

## Types of vectors of data

-   Logical vectors can take only three possible values:
    -   `FALSE`
    -   `TRUE`
    -   `NA` which is 'not available'.
-   Integers have one special value: NA, while doubles have four:
    -   `NA`
    -   `NaN` which is 'not a number'
    -   `Inf`
    -   `-Inf`

## Reading in Data Frames in R

-   A strength of `R` is being able to import data from an external source
-   Create the same table that you did above in a spreadsheet like Excel
-   Export it to comma separated and tab separated text files for importing into `R`.
-   The first will read in a comma-delimited file, whereas the second is a tab-delimited
-   In both cases the header and row.names arguments indicate that there is a header row and row label column
-   Note that the name of the file by itself will have R look in the CWD, whereas a full path can also be used

## Drawing samples from distributions

-   Here is a way to create your own data sets that are random samples.
-   Again, play around with the arguments in the parentheses to see what happens.

## Drawing samples from distributions

```{r}
#| label: fig-lec07-samples-from-distributions-1
#| fig-cap: "10,000 normal draws (x) plotted against 10,000 uniform draws (y)"
#| out-width: "100%"
#| echo: true
#| eval: true

x <- rnorm (10000, 0, 10)
y <- sample (1:10000, 10000, replace = T)
xy <- cbind(x,y)
plot(x,y) 
```

## Drawing samples from distributions

```{r}
#| label: fig-lec07-samples-from-distributions-2
#| fig-cap: "The same samples plotted from a two-column matrix made with cbind()"
#| out-width: "100%"
#| echo: true
#| eval: true

x <- rnorm (10000, 0, 10)
y <- sample (1:10000, 10000, replace = T)
xy <- cbind(x,y)
plot(xy)
```

## Drawing samples from distributions

```{r}
#| label: fig-lec07-samples-from-distributions-3
#| fig-cap: "Histogram of 10,000 draws from a normal distribution (mean 0, sd 10)"
#| out-width: "100%"
#| echo: true
#| eval: true

x <- rnorm (10000, 0, 10)
y <- sample (1:10000, 10000, replace = T)
xy <- cbind(x,y)
hist(x)
```

## Drawing samples from distributions

-   You’ve probably figured out that y from the last example is drawing numbers with equal probability.
-   What if you want to draw from a distribution?
-   Again, play around with the arguments in the parentheses to see what happens.

## Drawing samples from distributions

```{r}
#| label: fig-lec07-drawing-samples-from-distributions
#| fig-cap: "Histogram of normal draws with the theoretical density curve overlaid"
#| out-width: "100%"
#| echo: true
#| eval: true

x <-rnorm(1000, 0, 100)
hist(x, xlim = c(-500,500))
curve(50000*dnorm(x, 0, 100), xlim = c(-500,500), add=TRUE, col='Red')
```

-   `dnorm()` generates the probability density, which can be plotted using the `curve()` function.
-   Note that is curve is added to the plot using `add=TRUE`

## Visualizing Data

-   So far you've been visualizing just the list of output numbers
-   Except for the last example where I snuck in a `hist` function.
-   You can also visualize all of the variables that you've created using the `plot` function (as well as a number of more sophisticated plotting functions).
-   Each of these is called a `high level` plotting function, which sets the stage
-   `Low level` plotting functions will tweak the plots and make them beautiful

## Visualizing Data

-   What do you think that each of the arguments means for the plot function?
-   A cool thing about `R` is that the options for the arguments make sense.
-   Try adjusting an argument and see if it works
-   Note- most people use the package `GGPlot2` for nice scientific graphics

## Visualizing Data

```{r}
#| label: fig-lec07-visualizing-data
#| fig-cap: "A sequence from 0 to 10 plotted as points"
#| echo: true
#| eval: true

seq_1 <- seq(0.0, 10.0, by = 0.1) 
plot (seq_1, xlab="space", ylab ="function of space", type = "p", col = "red")
```

## Putting plots in a single figure

-   On the next slide
-   The first line of the lower script tells R that you are going to create a composite figure that has two rows and two columns. Can you tell how?
-   Now, modify the code to add two more variables and add one more row of two panels.

```{r}
#| label: lec07-putting-plots-in-a-single
#| out-width: "50%"
#| out-height: "20%"
#| echo: true
#| eval: true

seq_1 <- seq(0.0, 10.0, by = 0.1)
seq_2 <- seq(10.0, 0.0, by = -0.1)
```

## Putting plots in a single figure

```{r}
#| label: fig-lec07-composite-figure
#| fig-cap: "Four plots arranged in a 2 x 2 grid with par(mfrow)"
#| out-width: "50%"
#| out-height: "50%"
#| echo: true
#| eval: true

par(mfrow=c(2,2))
plot (seq_1, xlab="time", ylab ="p in population 1", type = "p", col = 'red')
plot (seq_2, xlab="time", ylab ="p in population 2", type = "p", col = 'green')
plot (seq_square, xlab="time", ylab ="p2 in population 2", type = "p", col = 'blue')
plot (seq_square_new, xlab="time", ylab ="p in population 1", type = "l", col = 'yellow')
```

## Creating Data Frames in R

-   As you have seen, in R you can generate your own random data set drawn from nearly any distribution very easily.
-   Often we will want to use collected data.
-   Now, let’s make a dummy dataset to get used to dealing with data frames
-   Set up three variables (hydrogel_concentration, compression and conductivity) as vectors

```{r}
#| label: lec07-creating-data-frames-in-r
#| echo: true
#| eval: true

hydrogel_concentration <- factor(c("low", "high", "high", "high", "medium", "medium", "medium","low"))
compression <- c(3.4, 3.4, 8.4, 3, 5.6, 8.1, 8.3, 4.5)
conductivity <- c(0, 9.2, 3.8, 5, 5.6, 4.1, 7.1, 5.3)
```

-   Create a data frame where vectors become columns

```{r}
#| label: lec07-creating-data-frames-in-r-2
#| echo: true
#| eval: true

mydata <- data.frame(hydrogel_concentration, compression, conductivity)
row.names(mydata) <- c("Sample_1", "Sample_2", "Sample_3", "Sample_4", 
                       "Sample_5", "Sample_6", "Sample_7", "Sample_8")
```

-   Now you have a hand-made data frame with row names
-   Take a look at it in the data section of RStudio

## Reading in and Exporting Data Frames

\

```{r}
#| label: lec07-reading-in-and-exporting-data
#| echo: true
#| eval: false

YourFile <- read.table('yourfile.csv', header=T, row.names=1, sep=',')
YourFile <- read.table('yourfile.txt', header=T, row.names=1, sep='\t')
```

\

```{r}
#| label: lec07-reading-in-and-exporting-data-2
#| echo: true
#| eval: false

write.table(YourFile, "yourfile.csv", quote=F, row.names=T, sep=",")
write.table(YourFile, "yourfile.txt", quote=F, row.names=T, sep="\t")
```

## Indexing in data frames

-   Next up - indexing just a subset of the data
-   This is a very important idea in R, that you can analyze just a subset of the data.
-   This is analyzing only the data in the file you made that has the factor value 'mixed'.

```{r}
#| label: lec07-indexing-in-data-frames
#| echo: true
#| eval: false

print (YourFile[,2])
print (YourFile$variable)
print (YourFile[2,])
plot (YourFile$variable1, YourFile$variable2)
```

## Practice reading in transcriptomic data

```{r}
#| label: lec07-practice-reading-in-transcriptomic-data
#| echo: true
#| eval: false

RNAseq_Data <- read.table('<name_of_file>', header=TRUE, sep=',')

print (RNAseq_Data)
head (RNAseq_Data)
tail (RNAseq_Data)

print (RNAseq_Data[,2])
print (RNAseq_Data[1,])
print (RNAseq_Data[1,2])
print (RNAseq_Data$ENSGACG00000000010)
print (RNAseq_Data$ENSGACG00000000010>45.0)
```

## Summary stats and figures

```{r}
#| label: lec07-summary-stats-and-figures
#| echo: true
#| eval: false

summary1 <- summary(RNAseq_Data $ENSGACG00000000003)
print (summary1)

hist(RNAseq_Data $ENSGACG00000000003)
boxplot(RNAseq_Data$ENSGACG00000000003)
boxplot(RNAseq_Data$ENSGACG00000000003~RNAseq_Data$Population)
plot(RNAseq_Data $ENSGACG00000000003, RNAseq_Data$ENSGACG00000000003)

boxplot(RNAseq_Data $ENSGACG00000000003~RNAseq_Data$Treatment, 
        col = "red", ylab = "Expression Level", xlab = "Treatment level", 
        border ="orange", 
        main = "Boxplot of variation in gene expression across microbiota treatments")
```

# Going further with R

## Installing vs. loading packages

``` r
install.packages("tidyverse")   # download & install - ONCE per computer
library(tidyverse)              # load - EVERY new R session
```

::: incremental
-   **`install.packages()`** puts a package on your computer (from CRAN)
-   **`library()`** makes it available in *this* session - restarting R means re-running `library()`
-   "there is no package called ..." means you never installed it
-   Bioconductor packages (genomics) use `BiocManager::install("name")`
-   Use one function without loading the package: `readxl::read_excel("file.xlsx")`
:::

## Inspecting objects and getting help

``` r
str(mydata)        # structure: type, dimensions, first values - your #1 tool
summary(mydata)    # quick summary of every column
class(mydata)      # what kind of object is this?
dim(mydata)        # rows x columns
names(mydata)      # column names
?mean              # help page for a function
??"linear model"   # search help across all packages
sessionInfo()      # R version + loaded package versions - put at the end of reports
```

::: callout-tip
When an object surprises you, run `str()` on it first. In **Positron** or **RStudio**, the Variables/Environment pane shows the same information.
:::

## Common gotchas that bite everyone

::: incremental
-   **R counts from 1**, and `x[-1]` *drops* the first element (it is not "the last" as in Python)
-   **`=` vs `==`**: `=` assigns or names an argument; `==` tests equality
-   **`NA` is contagious**: `mean(c(1, NA, 3))` is `NA` - use `na.rm = TRUE`
-   **Floating point**: `0.1 + 0.2 == 0.3` is `FALSE` - use `all.equal()`
-   **Case matters**: `Data` and `data` are different objects
-   **Working directory**: "cannot open file" usually means R is looking in the wrong folder - check `getwd()` and use project-relative paths
:::

## A taste of the tidyverse: the pipe `|>`

``` r
library(tidyverse)

mydata |>
  filter(compression > 4) |>                            # keep rows
  mutate(ratio = conductivity / compression) |>         # add a column
  group_by(hydrogel_concentration) |>                   # define groups
  summarise(n = n(), mean_ratio = mean(ratio))          # one row per group
```

| Verb          | What it does                       |
|---------------|------------------------------------|
| `filter()`    | Keep rows matching a condition     |
| `select()`    | Keep or drop columns               |
| `mutate()`    | Add or change columns              |
| `arrange()`   | Sort rows                          |
| `group_by()` + `summarise()` | Summaries per group |

::: callout-note
Read `|>` as **"and then"**. The older `%>%` pipe does the same thing.
:::

## A taste of ggplot2: the grammar of graphics

``` r
ggplot(mydata, aes(x = hydrogel_concentration, y = compression)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  labs(x = "Hydrogel concentration", y = "Compression") +
  theme_minimal()
```

::: incremental
-   **Data** + **aesthetic mappings** (`aes()`: which column goes to x, y, color...) + **geoms** (points, boxes, lines)
-   Layers are added with `+`
-   Same pattern for every plot type - learn it once
-   More in [Appendix D (R primer)](../Resources_Folder/Appendix_D_R_and_RStudio.qmd) and [Appendix E (R command reference)](../Resources_Folder/Appendix_E_Core_R_Commands.qmd)
:::
