---
title: "Week 6 Computational Tools for Bioengineers"
author: "Bill Cresko"
format: 
  revealjs:
    footer: BioE_Comp_Tools_2025 - Knight Campus 
    transition: fade
    transition-speed: slow
editor: visual
---

```{r}
library(tidyverse)
library(gt)
library(readxl)
theme_set(theme_minimal())
```

## Goals for today

-   Quarto Markdown
-   Vibe Coding with LLMs
-   Git and GitHub

# Markdown and Quarto Markdown

## What is Markdown? {.smaller}

::: {.incremental}
-   a very light weight but flexible editing language
-   allows the `rendering` of markdown text to a variety of different formats (pdf, html, web pages, powerpoint, etc....)
-   very simple and common across science and engineering
-   involves a `yaml` header (which stands for `yet another markdown language`)
-   similar to `html` but simpler
:::

## What is Quarto Markdown? {.smaller}

Quarto `(https://quarto.org/)` provides a next-generation authoring framework for data science.

-   `Quarto Markdown` is a powerful scientific and technical publishing system that extends the lightweight markup language Markdown for creating formatted text using a plain-text editor
-   You can use `Quarto Markdown` files to save and execute code \*and\* generate high quality reports that can be shared with an audience.
-   You can make html, pdf, docx, websites, slides, books, and more.

When you save your document, the output will be in the format indicated in your YAML header. This could be an html document, Word document, pdf, website, book, etc.

## Getting started {.smaller}

-   Open a new Quarto Markdown document from RStudio

-   You should see a document that is already populated with some example text and code

## This is Quarto Markdown {.smaller}

```{markdown, eval=F, echo=T}
---
title: "title"
output: html_document
---
```

-   Note that `---` delineates the `yaml header`.
-   This controls document-level settings and rendering.
-   There are a lot of options that you can put in the `yaml header`

## This is Quarto Markdown {.smaller}

```{markdown, eval=T, echo=T}
# Header size 1

## Header size 2

I can write text and it will look normal!

I can still run code!
```

## This is Quarto Markdown {.smaller}

```{r, eval=FALSE, echo=TRUE}
# ```{r code_chunk_name option1 option 2}
# x <- 5
# print(x^2)
# ```
```

-   By default, the code and output will print to your document
-   Note the back tick marks - these delineate the code
-   The {r} indicates this is r code

## Navigating chunks {.smaller}

-   Easy to run chunks - a few ways to do this
-   Top right of each chunk - just run, run all above, etc.
-   Cog wheel at the top of the Quarto Markdown document - run options
-   Keyboard shortcuts

## Chunk options {.smaller}

-   `include = FALSE` prevents code and results from appearing in the finished file. R Markdown still runs the code in the chunk, and the results can be used by other chunks.
-   `echo = FALSE` prevents code, but not the results from appearing in the finished file. This is a useful way to embed figures.
-   `message = FALSE` prevents messages that are generated by code from appearing in the finished file.
-   `warning = FALSE` prevents warnings that are generated by code from appearing in the finished.
-   `fig.cap = "..."` adds a caption to graphical results.
-   `eval = FALSE` shows code, but doesn't run it
-   `results = 'hide'` runs code, but doesn't show results
-   `fig.height, fig.width` set the dimensions of plots
-   `cache = TRUE` saves the results of a code chunk, can be helpful if a particular code chunk is intensive

## Quarto Markdown formatting {.smaller}

``` markdown
*italic*
**bold**
$A = \pi \times r^{2}$
```

*italic*

**bold**

$A = \pi \times r^{2}$

## Quarto Markdown lists {.smaller}

``` markdown
* Item 1
* Item 2
    + Item 2a
    + Item 2b
```

-   Item 1
-   Item 2
    -   Item 2a
    -   Item 2b

## Quarto Markdown lists {.smaller}

``` markdown
1. Item 1
2. Item 2
3. Item 3
    + Item 3a
    + Item 3b
```

1.  Item 1
2.  Item 2
3.  Item 3
    -   Item 3a
    -   Item 3b

## Typesetting Mathematics

-   Since 'tex' is embedded in Quarto Markdown you can typeset mathematics easily using LaTeX (https://en.wikibooks.org/wiki/LaTeX)

-   Essentially, just surround your mathematics in \$ symbols

-   \$\$ centers the expressions

## Typeset mathematics

Poisson Sampling Equation

```{r, eval=FALSE, echo=TRUE}
$$\large Pr(Y=r) = \frac{e^{-\mu}\mu^r}{r!}$$
```

$$\large Pr(Y=r) = \frac{e^{-\mu}\mu^r}{r!}$$

## Differential Equations {.smaller}

```{r, eval=FALSE, echo=TRUE}
$$\int xy^2\,dx\,dy =\frac{1}{6}x^2y^3$$
```

$$\iint xy^2\,dx\,dy =\frac{1}{6}x^2y^3$$

## In-line versus fenced {.smaller}

```{r, eval=FALSE, echo=TRUE}
This equation, $y=\frac{1}{2}$, is included inline
```

This equation, $y=\frac{1}{2}$, is included inline

```{r, eval=FALSE, echo=TRUE}
Whereas this equation, $$y=\frac{1}{2}$$, is put on a separate line
```

Whereas this equation $$y=\frac{1}{2}$$ is put on a separate line

## Matrix formulations {.smaller}

```{r, eval=FALSE, echo=TRUE}
$$	\begin{matrix}
		-2 & 1 & 0 & 0 & \cdots & 0  \\
		1 & -2 & 1 & 0 & \cdots & 0  \\
		0 & 1 & -2 & 1 & \cdots & 0  \\
		0 & 0 & 1 & -2 & \ddots & \vdots \\
		\vdots & \vdots & \vdots & \ddots & \ddots & 1  \\
		0 & 0 & 0 & \cdots & 1 & -2
	\end{matrix} $$
```

$$  \begin{matrix}
        -2 & 1 & 0 & 0 & \cdots & 0  \\
        1 & -2 & 1 & 0 & \cdots & 0  \\
        0 & 1 & -2 & 1 & \cdots & 0  \\
        0 & 0 & 1 & -2 & \ddots & \vdots \\
        \vdots & \vdots & \vdots & \ddots & \ddots & 1  \\
        0 & 0 & 0 & \cdots & 1 & -2
    \end{matrix} $$

## Quarto Markdown is very flexible {.smaller}

-   You can use Quarto templates and open as a new Quarto Markdown file
-   Better yet there are extensions that add functionality
    -   books
    -   journal articles
    -   slide shows (these slides!)
    -   interactive exercises

## More on Quarto Markdown {.smaller}

-   https://quarto.org/docs/authoring/
-   https://quarto.org/docs/reference/
-   Lots more fun stuff you can do!

------------------------------------------------------------------------

# `Vibe coding` with the ***help*** of Large Language Models (LLMs)

------------------------------------------------------------------------

# Git and GitHub

## What is Git? {.smaller}

-   **Git** is a distributed version control system
-   Tracks changes in your code over time
-   Allows collaboration with others
-   Maintains complete history of all changes
-   Essential tool for reproducible research
-   Works locally on your computer

## What is GitHub? {.smaller}

-   **GitHub** is a web-based hosting service for Git repositories
-   Provides cloud storage for your code
-   Enables collaboration features:
    -   Pull requests
    -   Issues tracking
    -   Project management
-   Hosts documentation and websites (GitHub Pages)
-   Free for public repositories

## Git Workflow Concept {.smaller}

```{mermaid}
%%| fig-width: 10
%%| fig-height: 6
flowchart LR
    A[Working Directory] -->|git add| B[Staging Area]
    B -->|git commit| C[Local Repository]
    C -->|git push| D[Remote Repository]
    D -->|git pull| A
    
    style A fill:#f9d71c
    style B fill:#87ceeb
    style C fill:#98fb98
    style D fill:#ffa07a
```

-   **Track Changes**: Every modification is recorded
-   **Collaboration**: Multiple people can work simultaneously
-   **Backup**: Your code is safe in multiple places
-   **Experimentation**: Try new things without breaking working code

## Setting Up Git {.smaller}

### 1. Install Git

``` bash
# Check if Git is installed
git --version

# Ubuntu/Debian
sudo apt-get install git

# macOS (with Homebrew)
brew install git

# Windows: Download from git-scm.com
```

### 2. Configure Git

``` bash
# Set your name and email
git config --global user.name "Your Name"
git config --global user.email "your.email@example.com"

# Check your settings
git config --list
```

# GitHub

## GitHub for Scientific Collaboration {.smaller}

### Perfect for Research Projects:

-   **Reproducibility**: Complete history of analysis development
-   **Collaboration**: Work with colleagues worldwide
-   **Documentation**: README files, wikis, and GitHub Pages
-   **Issues**: Track bugs, ideas, and to-dos
-   **Releases**: Tag specific versions for publications
-   **DOI Integration**: Link with Zenodo for citations

## Resources and Help {.smaller}

### Learning Resources:

-   GitHub Docs `(https://docs.github.com)`
-   Pro Git Book `(https://git-scm.com/book) (free online)`
-   GitHub Learning Lab `(https://github.com/apps/github-learning-lab)`
-   Quarto + GitHub Pages `(https://quarto.org/docs/publishing/github-pages.html)`

### Getting Help:

``` bash
# Git help
git help <command>
git <command> --help

# Quick reference
git help everyday
```

## Creating a GitHub Account {.smaller}

### Steps to Get Started:

1.  Go to [github.com](https://github.com)
2.  Click "Sign up"
3.  Choose a username (professional, memorable)
4.  Add your university email
5.  Verify your account
6.  Consider GitHub Student Developer Pack:
    -   Free GitHub Pro while you're a student
    -   Access to development tools
    -   Apply at [education.github.com](https://education.github.com)

## Basic Git Commands {.smaller}

### Initialize a Repository

``` bash
# Create a new repository
git init

# Clone an existing repository
git clone https://github.com/username/repository.git
```

### Track Changes

``` bash
# Check status
git status

# Add files to staging
git add filename.txt        # Add specific file
git add .                   # Add all changes

# Commit changes
git commit -m "Descriptive message about changes"
```

## Working with Remotes {.smaller}

``` bash
# Add a remote repository
git remote add origin https://github.com/username/repo.git

# View remotes
git remote -v

# Push changes to GitHub
git push origin main

# Pull changes from GitHub
git pull origin main

# Fetch without merging
git fetch origin
```

## Branching and Merging {.smaller}

```{mermaid}
%%| fig-width: 10
%%| fig-height: 6
gitGraph
    commit id: "Initial commit"
    commit id: "Add README"
    branch feature
    checkout feature
    commit id: "New feature"
    commit id: "Fix bug"
    checkout main
    commit id: "Update docs"
    merge feature
    commit id: "Continue work"
```

``` bash
# Create and switch to new branch
git checkout -b feature-branch

# Switch branches
git checkout main

# Merge branch
git merge feature-branch
```

## Common Git Workflow {.smaller}

### Daily Development Cycle

``` bash
# 1. Start your day - update your local copy
git pull origin main

# 2. Create a feature branch
git checkout -b new-feature

# 3. Make changes and commit regularly
git add .
git commit -m "Add new analysis function"

# 4. Push your branch
git push origin new-feature

# 5. Create pull request on GitHub (via website)
```

## Viewing History {.smaller}

``` bash
# View commit history
git log

# Prettier one-line format
git log --oneline

# With graph
git log --graph --oneline --all

# Show changes in commits
git log -p

# Show last n commits
git log -3
```

## GitHub Pages {.smaller}

### Host Your Research Website for Free!

```{mermaid}
%%| fig-width: 10
%%| fig-height: 5
flowchart LR
    A[Quarto Documents] -->|Render| B[HTML Files]
    B -->|git push| C[GitHub Repository]
    C -->|Automatic| D[GitHub Pages Website]
    D --> E[yourname.github.io/project]
    
    style A fill:#e3f2fd
    style E fill:#c8e6c9
```

### Setup Steps:

1.  Create repository
2.  Push your HTML files
3.  Enable Pages in repository settings
4.  Your site is live at `https://username.github.io`

## GitHub Pages Configuration {.smaller}

### In your repository:

1.  Go to Settings → Pages
2.  Source: Deploy from a branch
3.  Branch: main (or gh-pages)
4.  Folder: / (root) or /docs

### For Quarto projects:

``` yaml
# In _quarto.yml
project:
  type: website
  output-dir: docs

publish:
  gh-pages:
    branch: gh-pages
```

``` bash
# Publish with Quarto
quarto publish gh-pages
```

## Best Practices {.smaller}

### Commit Messages

-   Be descriptive and concise
-   Use present tense: "Add feature" not "Added feature"
-   Reference issues: "Fix bug #123"

### Repository Organization

```         
project/
├── README.md           # Project description
├── data/              # Raw data (consider .gitignore)
├── analysis/          # Analysis scripts
├── results/           # Output files
└── docs/              # Documentation
```

## .gitignore File {.smaller}

### Exclude files from version control:

``` bash
# Create .gitignore file
touch .gitignore
```

### Common patterns to exclude:

``` bash
# Data files
*.fastq
*.fasta
*.bam
*.vcf
data/raw/*

# R files
.Rhistory
.RData
.Rproj.user

# Output files
results/temp/*
*.log

# System files
.DS_Store
```
